---
title: "Foundational Topics in Measurement"
author: "Michael Woller"
draft: true
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: false
  warning: false
  message: false
standalone: true
---

## Introduction

A foundational part of psychological science is defining and measuring what we study. Before we can examine the causes of depression, test whether a treatment works, or study individual differences in personality, we need instruments that accurately capture the constructs we care about. If this sounds straightforward reframe how you understand people. Psychological constructs like depression, anxiety, impulsivity, or well-being do not permit direct observation. You can measure someone's height directly. You cannot measure their conscientiousness the same way that is universally consistent.

This creates a problem that no amount of methodological rigor downstream can fix. If your construct doesn't actually capture what you want it to, then there is nothing you can do to improve your inference into that construct.

If your depression scale is not measuring depression, a perfect model will stil produce meaningless results.

Flake and Fried (2020)^1^ coined the term **questionable measurement practices** (QMPs) to describe decisions researchers make that raise doubts about the validity of their measures and the validity of their conclusions. These include

-   failing to define the construct being measured

-   selecting measures without justification

-   modifying existing scales without reporting it

-   creating measures on the fly with no validity evidence.

Their review of the published literature found that between 40% and 93% of measures used across several journals lacked validity evidence, and that 69% of emotion measurement instances included no reference to prior research or systematic development. There are over 280 different scales for depression!

The goal of this course is to give you the conceptual and technical tools to do better. That means understanding what constructs are and how they relate to the things we observe, how to build and evaluate instruments that measure them accurately, and how to make and justify the many decisions involved in measurement transparently.

This page covers three foundational topics that underlie everything else we do in psychometrics:

1)  how to think about constructs and their hierarchical structure

2)  the mathematics of path analysis and how it connects observed correlations to underlying causal structure

3)  the major families of latent variable models and the critical distinction between reflective and formative measurement.

But before we can talk about reliability, validity, factor analysis, or item response theory, we need a shared vocabulary for what a psychological measure actually *is*. Psychological constructs (like depression, wellbeing, mathematical competency) cannot be observed directly. Instead, we observe **indicators**: test item responses, behaviors, or scores that we believe reflect some underlying construct. In brief, constructs are things we cannot measure but want to know. The "indicate" the values of the constructs. Indicators are things we can observe and measure in their stead.

The central questions of measurement are: what is the relationship between those indicators and the construct, and how should we model it?

We will be using the Depression Anxiety Stress Scales provided by the Open Psychometrics Project from the `networktree` package^2^. The dataset is named `dass`.

The **Depression Anxiety Stress Scales (DASS)** will serve as our running example throughout this page. It is a 42-item self-report instrument that asks respondents to rate how much each statement applied to them over the past week on a four-point scale (0 = did not apply, 3 = applied very much). Items are grouped into three 14-item subscales: depression, anxiety, and stress. Each of these items targets a distinct but related aspect of negative affect.

------------------------------------------------------------------------

## 1. Constructs and Hierarchical Analysis

Psychological constructs exist at different levels of breadth. Broad traits like neuroticism or conscientiousness sit at the top of a hierarchy. Narrow, domain-specific construct, like fear of negative evaluation, or anxiety specifically about statistics, sit lower. These are considered **latent variables** since they are not observed, but are presumed to still exist. They are "latent" in that they are unobservable.

```{r setup, include = F}
library(networktree)
library(networktree)
library(dplyr)
library(lavaan)
library(semPlot)
data(dass)
```

The DASS scale has the three following sub scales: Depression, Anxiety, Stress. The DASS is built on the assumption that depression, anxiety, and stress represent distinct but related negative emotional states. Depression reflects low positive affect, hopelessness, and anhedonia. Anxiety reflects physiological arousal and fear. Stress reflects chronic tension, irritability, and difficulty relaxing.

| Subscale | Example Items |
|------------------------------|------------------------------------------|
| Depression | "I couldn't seem to experience any positive feeling at all" |
|  | "I felt that life was meaningless" |
| Anxiety | "I felt scared without any good reason" |
|  | "I was aware of the action of my heart in the absence of physical exertion" |
| Stress | "I found it hard to wind down" |
|  | "I tended to over-react to situations" |

The quesiton remains on whether these items are distinct and are measuring different constructs or how much overlap is there between the constructs per items are there. This is exactly the kind of question hierarchical analysis is designed to answer.

Consider that the DASS depression items focus on low positive affect and hopelessness. The anxiety items focus on physiological arousal and fear. The stress items focus on tension, irritability, and difficulty relaxing. These are related but conceptually distinguishable. You could imagine a patient who scores high on stress but not depression. The question is whether that distinction shows up in the data, and that is the point of validating your measurement scale.

The `dass` data is made up of 58 columns of variables with a sample of 5,000. We can get a peak by using `glimpse()` from the `dplyr` package

```{r visual}
glimpse(dass)
```

We can see at the top of the `dass` dataset the 42 items associated with the 3 scale factors (note, "factor" is often used in place of construct, as they both often refer to latent variables). The label "A", "D", and "S" at the end of the item represent if that item is built to measure anxiety, depression, and stress respectively. So, when an individual takes this test, 14 items measure depression, 14 items measure anxiety, and 14 items measure stress. The other output on this datasetare data collected outside the DASS.

------------------------------------------------------------------------

## 2. Path Analysis

Before we can get into the weeds of how items measure constructs, we need to go over some basics of path analysis. Path analysis is a way of drawing and quantifying relationships among variables. It lets us express ideas like "X predicts Y" or "X and Y share a common source of variation" in a precise mathematical language. This matters for psychometrics because measurement models are special cases of path models. Understanding how to read and trace a path diagram will make everything that follows much easier to interpret.

### How to Read a Path Diagram

Here is an example of a path diagram for the Anxiety construct in the `dass` data using the `semPlot` package. We will cover how this R code works more in the future, so don't worry about it for now.

```{r, echo = T}
# Fit a model by expressing the construct 
model_anxiety <- '
  anxiety =~ dass2_A + dass4_A + dass7_A + dass9_A + dass15_A +
             dass19_A + dass20_A + dass23_A + dass25_A + dass28_A +
             dass30_A + dass36_A + dass40_A + dass41_A
'

fit_anxiety <- cfa(model_anxiety, data = dass)

semPaths(fit_anxiety,
         what        = "std",
         layout      = "tree",
         residuals   = TRUE,
         intercepts  = FALSE,
         nCharNodes  = 6,
         edge.color  = "black",      # all edges black
         fade        = FALSE,        # no opacity scaling by magnitude
          fixedStyle  = 1,            # force all lines solid (no dashes)
         edge.width  = .5,            # thin uniform line width
         asize       = 3,            # smaller arrowheads
         edge.label.cex = 0.6)       # shrink loading labels a bit   
```

Path diagrams like this are built from a small set of conventions:

-   **Squares** represent observed variables. These are the things we directly measure, like a person's response to a single DASS item.
    -   The squares are the 14 items that measure an individual's anxiety. E.g., their response to ""I felt scared without any good reason". (The number corresponds to the number in order this item appears the in the list of 42 questions).
-   **Circles** represent latent variables. These are constructs we cannot directly measure, like anxiety or depression (`semPlot` abbreviates the names, hence why it says "anxity").
-   **Straight single-headed arrows** represent a directional relationship, where the variable at the tail predicts the variable at the head. It represents a regression model with the outcome ast the end of the arrow. The direction of the arrow matters: it denotes the direction that the variables influence each other (we will come back to this).
-   **Curved double-headed arrows** between two variables represent a correlation (not present in this model). This means the variables are related but neither is specified as the cause of the other.
-   **Small arrows or loops** pointing into a variable from outside represent its residual (we see the looped arrow here), which is the variance not explained by anything else in the model. For observed items these are called residuals or measurement error (notated as $\varepsilon$). For latent variables they are called disturbances (notated as $\zeta$).
-   **Numbers on the arrows** are estimated values from the data. When it is on curved arrows, these are the values of the correlations/residual variances. When the number is on and arrow that runs from a latent variable to an observed item, this coefficient is called a "factor loading".

In the path diagram above, we can see that the latent factor of "anxiety" predicts how an individual responds to the items on the DASS associated with anxiety. We know this model is saying anxiety predicts the items since the arrows are pointed in that direction. This is called a **reflective model** -- the construct causes the items, not the other way around. We will formalize this distinction in the next section.

A core idea with path diagrams is that the association between any two variables can be understood by following a path of lines. If you wish to figure out this association you can trace all valid routes connecting them, multiply the coefficients along each route, and sum the results. There are three general caveats however:

1.  **No loops.** In tracing from one variable to another, you cannot pass through the same variable twice on a single route.

2.  **No going forward then backward.** Once you have traveled forward along a directed path, you cannot reverse direction to reach the endpoint.

3.  **Only one curved arrow per route.** A curved double-headed arrow represents a pre-existing correlation between two variables. A traced route can traverse at most one of these.

If you follow these rules, you can deduce the correlation between any two variables by multiplying the coefficients along each route and sum across the routes.

### A Tracing Example

Looking at the diagram above, the latent factor **anxiety** predicts each item with a standardized loading ("loading" since the arrow comes from a circle latent variable to a square observed variable, and "standardized" since its loaded from a standardized anxiety construct, which was just something I chose to do arbitrarily for this lesson). Using the tracing rules, we can express the implied correlation between any two items as the product of their loadings through their shared latent variable anxiety. For example the correlation between the 2nd and 4th DASS question items (which both measure anxiety) is:

$$r_{\text{dass2\_A},~\text{dass4\_A}} = 0.50 \times 0.70 = 0.35$$

That is the only valid route from `dass2_A` to anxiety, then back down to `dass4_A`.

A few more examples:

$$r_{\text{dass7\_A},~\text{dass25\_A}} = 0.71 \times 0.64 = 0.45$$

$$r_{\text{dass9\_A},~\text{dass28\_A}} = 0.68 \times 0.75 = 0.51$$

This generalizes to any pair of items in the model. We can formalize this mathematically. For items $i$ and $j$ with loadings $\lambda_i$ and $\lambda_j$. Their correlation is

$$r_{ij} = \lambda_i \times \lambda_j$$ $\lambda$ is the universal symbol for a "factor loading". Meaning if you ever need to state the relationship between the construct and items mathematically, it would be given as a $\lambda$.

------------------------------------------------------------------------

## 3. Reflective vs. Formative Measurement

So far, we've only talked about the arrow pointing one way: from the latent variable (circle) to the observed indicator variable (square). However, we can build the model two ways, pointing outwards from the latent variable like before or all the squares pointing *towards* the latent variable.

The distinction between the former **reflective** model (the items reflect the construct) and the latter **formative** (the items form the construct) is one that needs to be thought about carefully. Does your construct exist regardless of the item measurements? Then this would be best captured with a reflective mdoel. Presumably, "anxiety" would persist if we never measured it via the DASS, so we can use these items to reflect its latent nature. However, what if the construct only exists because of the items you've measured? Then this would ba a formative model as you are only creating the construct due to your measurements. An example of this would be *socioeconomic status (SES)*, which is typically measured by items like income, educational level, etc.

SES does not exist as an independent psychological state that causes you to have a certain income. There is also no reason its indicators need to correlate with each other. A self-employed tradesperson might have high income but low formal education. Dropping one of these indicators does not give you a noisier measure of the same construct. If we dropped one item from the DASS that measures anxiety, we would still be measuring the *same* anxiety, but it would just be a worse measure. However, if we dropped one of the indicators for SES, we would be measuring a completely new construct.

Figuring out what type of latent construct you have is the first step in creating a measurement model.

For example, you can think of **happiness** either through a reflective model or a formative model. If you see happiness as some kind of inherent positive affect an individual has, it makes sense to treat it as reflective as it's effect will permeate to measurable aspects. However, if you view happiness as constituted by life events—having a good job, having plenty of social activities each week, getting good rest—then you might view happiness through a formative model. How you treat your construct depends on your theory and reasoning for measuring it in the first place.

Some examples of constructs that Bollen and Lennox (1991)^3^ note have been misspecified as reflective when they should be formative include Life change, Illness, Monitoring capacity of an organization, and Instrumental activities of daily living. Personally, I've found old studies myself that incorrectly used reflective models (to the detriment of hundreds of clinical psychologists using their wrong model decades down the line).

### Reflective Models

In a reflective model, the latent variable *causes* the indicators. Therefore, the observed items are the effect of the latent variable (cause $\rightarrow$ effect). Thus, the indicator variables of a reflective model are sometimes called *effect indicators*. If someone's underlying depression level increases, we expect all depression indicators to increase together, including sleep disturbance, sadness, hopelessness, and loss of appetite. Thiiss is because the common cause (depression) drives all of them. The indicators, as a result, should be highly intercorrelated, because they all reflect the same underlying state.

The mathematical relationship between a latent variable (commonly denoted as $\eta$) and the effect indicator measurement ($y$) is given as:

$$y_i = \lambda_i \eta + \varepsilon_i$$

In our example of the DASS items ($i$), the relationship between any item score $y_i$ (recall our indicator is a DASS test item) can be predicted by anxiety's ($\eta$) specific effect on that item ($\lambda_i$). However, our construct of anxiety won't explain every individual's score on that item perfectly, so we represent this discrepency via the measurement error for that specific item ($\varepsilon_i$).

This is the type of model that is the most of interest in psychometrics, and it will be the type of model that we will prediminatly focus on.

### Formative Models

In a formative model, the indicators *define* or *cause* the latent variable. So, in this context, the indicator variables are called **cause indicators**.

We talked about SES, but another example is **stressful life events**: being in a car accident, losing a job, experiencing a family illness, and relocating are not caused by some underlying "stress" latent variable. Rather, they *constitute* your total stress exposure. In this situation, there is no reason these indicators need to correlate with each other. You can have a car accident without losing your job. Therefore, the latent construct can be formed as just the summation of all of your indicators. This can be mathetmatically formalized as

$$\eta = \gamma_{1}x_1 + \gamma_{2}x_2 + \ldots + \gamma_{g}x_g + \zeta$$ Here, we are predicting stressful life events $\eta$ from each of our life events indicator's ($x$ up to the $g$'th indicator) effect on the construct ($\gamma$). Obviously, we can't account for *all* life events that could induce stress, the remaining unexplained aspect is given in the disturbance residual term ($\zeta$).

### Implications of the Distinction

**1. Internal consistency.** In a reflective model, the correlation between any two indicators from the same factor equals $\lambda_i \lambda_j$, the product of their loadings. High intercorrelations among items are expected and required. In a formative model, there is no such constraint. Indicators can be uncorrelated, because they have independent causes. This limits the types of validity checks that can be done on formative models

**2. Dropping indicators.** In a reflective model, dropping an indicator has little consequence. The remaining indicators still cover the same ground. In a formative model, dropping an indicator *changes the construct itself*. Every facet of the construct must be represented.

**3. Linear composites as substitutes for latent variables.** Researchers often form a sum or average score of items as a proxy for the latent variable (e.g., summing all the anxiety items and stating this score is your "anxiety level"). This is never equivalent to the latent variable, regardless of the model type. In a reflective model, the composite carries measurement error $\varepsilon$ from the individual items. In a formative model, the composite carries the equation disturbance $\zeta$, which is uncorrelated with the $x$'s. Eitherway, both of the models will become filled with error.

Concretely with an example, if anxiety truly correlates .60 with some health outcome, but you measure anxiety with a sum score of test items instead of a latent variable, your observed correlation might only come out to .45. The true relationship is being **attenuated** by measurement error (aka, you are measuring the relationship poorly and are missing the mark).

### The Ongoing Debate

The reflective/formative distinction has generated substantial methodological debate. Edwards (2011)^4^, in "The Fallacy of Formative Measurement," argues that the enthusiasm for formative measurement in organizational research is based on conceptions of constructs and causality that are difficult to defend, and that the goals of formative measurement can usually be achieved through more involved reflective specifications. Howell, Breivik, and Wilcox (2007)^5^ argue the opposite: that reflective measurement is often inappropriate for constructs that are genuinely defined by their indicators, and that formative models better represent the theoretical structure of many organizational constructs.

The practical takeaway is that the choice between reflective and formative measurement is a *theoretical* choice, not a statistical one. It depends on whether the construct exists conceptually prior to its indicators (reflective) or constructed from them (formative). Making the choice deliberately and evaluating the scale is what matters.

The DASS is a good example of a reflective instrument. Anxiety, depression, and stress are each treated as real underlying states that cause responses across their respective items. A person's anxiety level drives how they respond to all 14 anxiety items. The items don't define the anxiety, they reflect it. With that established, we can now ask a further question: how should the reflective structure among those 14 items for the three scales be organized?

------------------------------------------------------------------------

## 4. Latent Variable Measurement Models

There are many ways you can formulate a psychometric model. The four we will discuss are the four common methods to structure a psychological construct. The models we will be going over are reflective models. Where they differ is in how many latent variables are specified and how those latent variables relate to each other.

The DASS was built with the expectation that the 14 anxiety items behave similarly. Someone who scores high on `dass25_A` should also score high on `dass7_A`. They share covariance because they share a common cause—the latent anxiety construct. Someone high in anxiety should score high across all 14 items. Someone low in anxiety should score low across all of them. Following this logic, we can work backwards from the observed item responses to estimate where someone falls on the latent construct. Aka, based on how they score on the items, we can predict what are their latent anxiety levels that are causing their responses.

This is psychometrics in a nutshell, but this is easier said than done. The four model families below represent different theoretical claims about how that latent structure is organized.

We will cover how to construct the models in R later, for now, I want to cover their general forms.

### Unidimensional

The simplest case: all indicators load onto a single latent factor. This is appropriate when items are intended to measure one and only one underlying construct. E.g., anxiety items *only* predict anxiety. This matches what we've been doing above so far. The model is formalizd as

$$y_i = \lambda_i \eta + \varepsilon_i$$ Each observed score $y_i$ is a linear function (regression) of the latent variable $\eta$ (with loading $\lambda_i$) plus measurement error $\varepsilon_i$. *Items with larger loadings are better indicators of the construct.* To repreint the anxiety diagram here.

```{r, echo=FALSE}
# Fit a simple unidimensional CFA for anxiety items only
model_anxiety <- '
  anxiety =~ dass2_A + dass4_A + dass7_A + dass9_A + dass15_A +
             dass19_A + dass20_A + dass23_A + dass25_A + dass28_A +
             dass30_A + dass36_A + dass40_A + dass41_A
'

fit_anxiety <- cfa(model_anxiety, data = dass)

semPaths(fit_anxiety,
         what        = "std",
         layout      = "tree",
         residuals   = TRUE,
         intercepts  = FALSE,
         nCharNodes  = 6,
         edge.color  = "black",      # all edges black
         fade        = FALSE,        # no opacity scaling by magnitude
          fixedStyle  = 1,            # force all lines solid (no dashes)
         edge.width  = .5,            # thin uniform line width
         asize       = 3,            # smaller arrowheads
         edge.label.cex = 0.6)       # shrink loading labels a bit   
```

Looking at this plot, we cans see that `dass28_A` has the highest loading at .75. If we check the code book of what each item is, our model says the best question at measuring anxiety was "*I felt I was close to panic*". This makes sense intuitively since panic is an accute expression of anxiety.

We can also get this from the `summary()` function. Since we want to look at standardized loadings for this example, add `standardized = TRUE` for the output to include these.

```{r}
summary(fit_anxiety, standardized = TRUE)
```

Here in the `Std.all` column of the "Latent Variables" table, we can see the standardized loadings, along with the unstandardized loadings and their z-tests, which can help determine if these are significantly different than 0 or not (aka, they help measure the construct or not). At the bottom, we can see the variances for every item as well as the latent construct. Recall, these variances represent the ammount of variation that is not explained by the model. So `dass2_A` has the most variance of individual's responses that is not explained by the anxiety construct.

### Correlated Factors

Multiple latent factors, each measured by its own set of indicators, with the factors allowed to correlate with each other. This is the standard confirmatory factor analysis (CFA) model for multidimensional constructs (we will learn more about 'CFAs' in the future). The correlation between two indicators from different factors is explained by the correlation between those factors.

```{r, echo =T}
# Fit a correlated CFA for each construct that are formed from their respective 14 items
model_correlated <- '
  depression =~ dass3_D + dass5_D + dass10_D + dass13_D + dass16_D +
                dass17_D + dass21_D + dass24_D + dass26_D + dass31_D +
                dass34_D + dass37_D + dass38_D + dass42_D
  anxiety    =~ dass2_A + dass4_A + dass7_A + dass9_A + dass15_A +
                dass19_A + dass20_A + dass23_A + dass25_A + dass28_A +
                dass30_A + dass36_A + dass40_A + dass41_A
  stress     =~ dass1_S + dass6_S + dass8_S + dass11_S + dass12_S +
                dass14_S + dass18_S + dass22_S + dass27_S + dass29_S +
                dass32_S + dass33_S + dass35_S + dass39_S
'

fit_correlated <- cfa(model_correlated, data = dass)

semPaths(fit_correlated,
         what        = "std",
         layout      = "tree",
         residuals   = F,        # drop residuals to reduce clutter
         intercepts  = F,
         nCharNodes  = 6,
         edge.color  = "black",
         fade        = FALSE,
         lty         = 1,
         edge.width  = 0.3,
          fixedStyle  = 1,
         asize       = 3,
         edge.label.cex = .1,         # hide loading labels entirely
         sizeMan     = 3,            # shrink item boxes
         sizeLat     = 10)
```

This is what the path diagram would look like with the 3 latent factors loading onto their 14 indicator items. Notice the curved arrows between the factors. This shows why it is a "correlated" model. The factors do *not* exist independently, which intuitively makes sense. As someone's stress is probably related to their depression or anxiety to some degree. This figure does not include the variances (the circle arrows) nor the specific values of the loadings and correlations due to simplicity. When making path diagrams, the labels can get clutered very quickly, as you can see with the overlapping square items boxes. It's best to use dedicated programs to make these figures.

I can draw out the specific values using the `summary()` function. Since we want to look at standardized loadings for this example, add `standardized = TRUE` for the output to include these.

```{r}
summary(fit_correlated, standardized = TRUE)
```

Here I can see that the factor loading for each factor in the top table labeled "Latent Variables". Since we are standardized our model to make the scale simpler, we should look at the last column that says `Std.all`.

We can see that the highest standardized loading for depression was `dass21_D` with .837, which corresponds to the question "I felt that life wasn't worthwhile".

The highest loading for stress was `dass11_S` with .739, which corresponds to the question "I found myself getting upset rather easily".

Lastly, the highest loading for anxiety was `dass20_A` with .739, which corresponds to the question "I felt scared without any good reason". It is noteworthy that the highest loading for anxiety changed from the unidimensional model, where `dass28_A` ("I felt I was close to panic") had the highest loading at .75. So, allowing anxiety to correlate with depression and stress shifted which item best represents the construct, albeit only slightly. This is a small but meaningful illustration of why model choice matters: the loadings you obtain and your interpretation of what best defines a construct depends on the broader structure you impose on the data. An item that looks like the best indicator of anxiety in isolation may not be the best indicator once you account for its overlap with depression and stress.

The bottom table, again, holds the variances of the items and constructs. However, an important new output is the "Covariances:" table that show the amount the constructs are related. If you look at the `Std.lv`, `Std.all` we can see the correlations between the factors (`Std.lv` is the covariances when only the latent factors are standardized and `Std.all` is the covariances when every variable is standardized. Note, these will be the same for the latent factors since the latent factors are standardized in both). Recall that a standardized covariance is a correlation, hence why we look at these columns for the correlations.

But here we can see that the correlations between each of the factors is very high. The correlation between anxiety and depression is high with .712, but the other correlations are even higher with depression and stress with a correlation of .797 as well as anxiety and stress with a correlation of .873. So it seems it was wise to consider these constructs as non-independent. But now we can ask the question of what might explain this relationship between what we observe.

### Second-Order (Hierarchical)

A higher-order latent factor causes several lower-order factors, which in turn cause the observed indicators. This model is appropriate when you believe the subscale factors are themselves explained by a broader overarching construct. You can have any number of "higher order" factors, though it is often hard to validate anything beyond a second-order (two factor layers). These models are sometimes refered to as hierarchical factor models as there is a heiarchy of factors.

Aka, there is another latent variable construct that is causing depression, anxiety, and stress. This would act as a potential explanation for why these constructs are correlated. An example for what this might be in DASS would be "negative affect". That is, someone high in negative affect would have higher levels on all three constructs So it actually isn't of interest to look at these three constructs in specific. Rather, as researchers we should be more intereted in the validity of a potential negative affect factor.

For another example of where a higher-order construct might occue is with a general Disability factor might cause specific factors for Household functioning, Cognitive functioning, Mobility, Self-care, Social functioning, and Society participation, with each of those factors measured by its own items. The second-order factor (Disability) explains why the first-order factors correlate with each other.

```{r, echo = T}
# form the second-order factor by  making the model for the first-order constructs
# then make the second-order construct from the first-order constructs
model_secondorder <- '
  depression =~ dass3_D + dass5_D + dass10_D + dass13_D + dass16_D +
                dass17_D + dass21_D + dass24_D + dass26_D + dass31_D +
                dass34_D + dass37_D + dass38_D + dass42_D
  anxiety    =~ dass2_A + dass4_A + dass7_A + dass9_A + dass15_A +
                dass19_A + dass20_A + dass23_A + dass25_A + dass28_A +
                dass30_A + dass36_A + dass40_A + dass41_A
  stress     =~ dass1_S + dass6_S + dass8_S + dass11_S + dass12_S +
                dass14_S + dass18_S + dass22_S + dass27_S + dass29_S +
                dass32_S + dass33_S + dass35_S + dass39_S
  negaffect  =~ depression + anxiety + stress
'

fit_secondorder <- cfa(model_secondorder, data = dass)

semPaths(fit_secondorder,
         what        = "std",
         layout      = "tree",
         residuals   = TRUE,
         intercepts  = FALSE,
         nCharNodes  = 6,
         edge.color  = "black",
         fade        = FALSE,
         fixedStyle  = 1,
         edge.width  = .5,
         asize       = 3,
         edge.label.cex = 0.1)
```

As you might have noticed, the diagram gets even messier to read. This the point where diagrams might start abbreviating the number of items they have by only depicting a few in a way where we can infer the rest. What is important to note here is that there is no longer any correlations between the 3 constructs (no curved arrows between them). Instead, their relationships are 100% determined by the second-order factor negative affect.

Eitherway, we can get the full results by using `summary()` with `standardized = TRUE` once again.

```{r}
summary(fit_secondorder, standardized = TRUE)
```

I won't go over all the loadings and variances again, but you might notice that some of the relationships have changed once again. This is because the inclusion of the second-order negative affect factor changes what each first-order factor needs to explain. The shared variance among depression, anxiety, and stress is now attributed to the second-order factor above them rather than being estimated as direct correlations between the subscales (the three original factors). This means the second-order model is making a stronger theoretical claim: that the reason the three subscales correlate is not just that they happen to co-occur, but that a single underlying dimension of general negative affect drives all three. If that claim is justified, the model is more parsimonious (i.e. simpler and better). If it is not justified, the correlated factors model is the more honest representation of the data. We will talk more about how to validate these in the future.

The new and important output to turn to is the output following `negaffect =~`, which describes the relationship between negative affect and each of the three first-order factors. Turn to the `Std.all` column. The loading of depression onto negative affect is .806, anxiety is .883, and stress is .988. This tells us that stress is almost entirely explained by the general negative affect factor (it is almost at the maximum of 1, so there is very little variance in stress that is not accounted for by general negative affect). Anxiety and depression retain somewhat more unique variance, with depression being the most distinct of the three at .806. This pattern is consistent with what we saw in the correlated factors model, where anxiety and stress had the highest interfactor correlation (.873). The second-order model is now expressing that same near-redundancy through the stress loading on the general factor. It raises a reasonable question about whether stress is doing enough unique work to justify being treated as a separate construct from the second-order negative affect construct at all.

The value of these higher order loadings can give you some insight to how valid this structure is. After all, if the second-order factor has a very small loading factor, it is probably not very good at explaining the subscales. However, this isn't the best method of validating the model structures. We will cover how to do this in the future.

### Bifactor

In this type of model, each indicator loads on both a **general factor** (shared by all items regardless of construct) and one **specific factor** (shared only within that item's construct). Aka, each anxiety, depression, and stress item load onto their constructs like normal. However, the *also* all load onto a seperate general factor together.

Unlike the second-order model, the general factor directly explains item variance alongside the specific factors. The general and specific factors are constrained to be uncorrelated with each other.

This model is particularly useful for constructs like psychopathology, where items might reflect both a broad general internalizing/externalizing issue and also reflect a more specific symptom clusters (depression, anxiety, obsessive-compulsive symptoms, etc.).

Here are some popular psychological constructs often constructed as a bifactor model:

**Psychopathology / Mental Health** The *p* factor, a general psychopathology factor underlying all mental disorder. There's many specific variations of this.

**Intelligence** The *g* factor (general intelligence) with specific factors for verbal, spatial, processing speed, and working memory is the classic historical example. However, there are many other intelligence measures that use different subscale factors.

**Personality** The Big Five has been examined as a bifactor model with a general factor sometimes called the Big One or General Factor of Personality (GFP), though as far as I can tell, this is controversial (bifactor models are often very controversial regardless.

**Depression specifically** The PHQ-9 and BDI-II have both been examined as bifactor models with a general depression factor and specific factors for somatic symptoms, cognitive symptoms, and affective symptoms that manifest within depression.

For the DASS, a bifactor model would say that each of the 42 items is driven by two things simultaneously: a general negative affect factor shared by all items, and a specific factor unique to the item subscales (depression-specific, anxiety-specific, or stress-specific). This is a different claim from the second-order model. In the second-order model, the general factor exists at a high-order and only influences items *indirectly* through the first-order factors. In the bifactor model, the general factor hits every item *directly*, and the specific factors capture what is left over after that general variance is removed.

Put another way, the bifactor model explains each individual's item responses from the general negative affect factor, and whatever variance of responses that remain unexplained by that general factor is then attributed to the specific subscale construct.

```{r bifactor}
# For a bifactor model, fit individual subscales with their specific items
# and also fit a general factor that is made up of every single item combined
# lastly make sure none of the factors are correlated, or make them "orthogonal"
# more detail on this type of code will be given in the future.
model_bifactor <- '
  # General factor loads on all 42 items
  negaffect    =~ dass3_D + dass5_D + dass10_D + dass13_D + dass16_D +
                dass17_D + dass21_D + dass24_D + dass26_D + dass31_D +
                dass34_D + dass37_D + dass38_D + dass42_D +
                dass2_A + dass4_A + dass7_A + dass9_A + dass15_A +
                dass19_A + dass20_A + dass23_A + dass25_A + dass28_A +
                dass30_A + dass36_A + dass40_A + dass41_A +
                dass1_S + dass6_S + dass8_S + dass11_S + dass12_S +
                dass14_S + dass18_S + dass22_S + dass27_S + dass29_S +
                dass32_S + dass33_S + dass35_S + dass39_S
  # Specific factors
  depression =~ dass3_D + dass5_D + dass10_D + dass13_D + dass16_D +
                dass17_D + dass21_D + dass24_D + dass26_D + dass31_D +
                dass34_D + dass37_D + dass38_D + dass42_D
  anxiety    =~ dass2_A + dass4_A + dass7_A + dass9_A + dass15_A +
                dass19_A + dass20_A + dass23_A + dass25_A + dass28_A +
                dass30_A + dass36_A + dass40_A + dass41_A
  stress     =~ dass1_S + dass6_S + dass8_S + dass11_S + dass12_S +
                dass14_S + dass18_S + dass22_S + dass27_S + dass29_S +
                dass32_S + dass33_S + dass35_S + dass39_S
  # Constrain all factors to be orthogonal
  negaffect    ~~ 0*depression
  negaffect    ~~ 0*anxiety
  negaffect    ~~ 0*stress
  depression ~~ 0*anxiety
  depression ~~ 0*stress
  anxiety    ~~ 0*stress
'

fit_bifactor <- cfa(model_bifactor, data = dass)

semPaths(fit_bifactor,
         what        = "std",
         layout      = "tree2",    # put tree2 or tree3 to make the diagram the correct style
         bifactor    = "negaffect",# make sure this is the name of your general factor
         residuals   = FALSE,
         intercepts  = FALSE,
         nCharNodes  = 6,
         edge.color  = "black",
         fade        = FALSE,
         fixedStyle  = 1,
         edge.width  = .5,
         asize       = 3,
         edge.label.cex = 0.1)
```

The diagrams are getting even more messy that I removed the residual variances (circle arrows). It's actually pretty common to not put these circle arrows since they are usually implied to exist. But this would also be another example where researchers would start putting short hand for the number of items. Generally, for bifactor models it is appropriate to put the subscales on one side and the general factor on the other side, with the items either horizontal (like above) or vertical in the middle.

The key thing to look at in the output is whether the specific factors retain meaningful loadings after the general factor has been accounted for. If the specific factor loadings are near zero, it suggests the subscales are not capturing anything beyond general negative affect. In other words, the three-factor structure of the DASS is not adding much beyond a single dimension. If the specific loadings remain substantial, it supports the idea that depression, anxiety, and stress each capture something real and distinct over and above their shared negativity.

Run the summary to inspect both sets of loadings with `summary()` including `standardized = TRUE`:

```{r bifactor-summary}
summary(fit_bifactor, standardized = TRUE)
```

Look at the `Std.all` column for both the `general =~` rows and the specific depression, anxiety, and stress rows. Strong general loadings with weak specific loadings would suggest the DASS is essentially unidimensional for negative affect in practice. Strong loadings on both would support the bifactor structure and justify reporting subscale scores alongside a total score.

The general factor loadings are moderate to strong across all 42 items, ranging from .447 to .770, suggesting that general negative affect is a meaningful source of variance in every item regardless of which subscale it belongs to. However, the picture for the specific factors is more complicated.

The depression-specific factor retains respectable loadings, with most items in the .35-.61 range, suggesting that depression items capture something beyond general negative affect. So we could say that thereis a meaningful depression-specific dimension even after general negative affect is accounted for.

The anxiety-specific loadings tell a different story. Several items have very weak specific loadings: `dass9_A` (.098), `dass20_A` (.126), `dass30_A` (.053), `dass36_A` (.106), and `dass40_A` (.087) are all near zero, while only a few others like `dass7_A` (.555) and `dass41_A` (.522) remain meaningful. This suggests the anxiety subscale is not a coherent specific dimension in the bifactor sense. Some anxiety items are doing anxiety-specific work while others are essentially just indicators of general negative affect.

The examples of the qeustions that would be more negative affect than anxiety would be

`dass9_A` -- "I found myself in situations that made me so anxious I was most relieved when they ended"

`dass20_A` -- "I felt scared without any good reason"

`dass30_A` -- "I feared that I would be 'thrown' by some trivial but unfamiliar task"

`dass36_A` -- "I felt terrified"

`dass40_A` -- "I was worried about situations in which I might panic and make a fool of myself"

The stress-specific factor is the most troubled. Several items have negative specific loadings: `dass8_S` (-.056), `dass12_S` (-.183), and `dass33_S` (-.220), which means those items are actually negatively related to the stress-specific factor after the general factor is removed.

These items represent the questions

`dass8_S` -- "I found it difficult to relax"

`dass12_S` -- "I felt that I was using a lot of nervous energy"

`dass33_S` -- "I was in a state of nervous tension"

This is a red flag suggesting the stress-specific factor may not be a well-defined or coherent construct once general negative affect is already accounted for. This reflects what we saw from the second-order model, where negative affect loaded .988 (out of a maximum of 1) on stress, meaning that stress was almost completely just measuring negative affect. This raises a genuine question about whether stress is doing meaningful unique work as a construct in the DASS beyond what is already captured by general negative affect.

This is the benefit that a bifactor model can give. However, just because you get these types of results doesn't mean that a general factor is justified. To do that, we will need to examine many different validation techniques and statistics, which we will do in future lessons.

------------------------------------------------------------------------

## Well Done!

You have completed the Foundational Topics in Measurement page. Here is a summary of what was covered:

-   The concept of **questionable measurement practices (QMPs)** from Flake and Fried (2020) and why measurement decisions have real consequences for the validity of psychological research
-   The distinction between **constructs** (latent variables we cannot directly observe, like depression or anxiety) and **indicators** (observable items or scores that reflect those constructs)
-   The **DASS** as a running example: a 42-item self-report measure with three 14-item subscales for depression, anxiety, and stress, used throughout to illustrate measurement concepts with real data
-   How to read a **path diagram**: squares for observed variables, circles for latent variables, single-headed arrows for directional relationships, curved double-headed arrows for correlations, and small loops for residuals and disturbances
-   **Wright's tracing rules** for deriving implied correlations between variables: no loops, no going forward then backward, and only one curved arrow per route
-   The **reflective vs. formative** distinction: reflective models treat the construct as the cause of the indicators (effect indicators), while formative models treat the indicators as constituting the construct (cause indicators)
-   Why the reflective/formative choice is a theoretical decision, not a statistical one, and the practical consequences of getting it wrong: dropping an indicator changes the construct in a formative model but not a reflective one
-   The four major families of reflective latent variable models: **unidimensional** (all items load onto one factor), **correlated factors** (multiple factors with estimated correlations between them), **second-order** (a higher-order factor explains why the first-order factors correlate), and **bifactor** (each item loads on both a general factor and a specific factor simultaneously)
-   How to fit all four model families in R using `lavaan` and visualize them with `semPlot`, and how to interpret standardized factor loadings from `summary()` with `standardized = TRUE`
-   The applied findings from the DASS bifactor model: depression retains a meaningful specific factor after general negative affect is accounted for, anxiety shows mixed results with several items functioning as general rather than anxiety-specific indicators, and stress shows near-complete overlap with general negative affect, raising questions about whether it is doing unique construct work

The next page covers reliability theory, including classical reliability coefficients, omega, and bifactor reliability indices.
