---
title: "IRT Foundations"
author: "Michael Woller"
draft: true
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(ggplot2)
```

## Introduction

Imagine you are trying to measure something you cannot directly observe, like how anxious someone is, how much they agree with a political ideology, or how strong their reading ability is. You cannot just look at a person and read off their anxiety level the way you might read a thermometer. Instead, you ask them a set of questions, and you use their answers as a window into the underlying thing you actually care about. This is the core of psychological measurement as we've discussed before.

As we discussed previously, for most of the twentieth century, the dominant answer came from **classical test theory (CTT)**. CTT is intuitive: add up a person's item scores (questions on a test that have a point value), treat the total as their measurement, and use statistics like Cronbach's alpha to tell you how much random noise is contaminating the signal. It works, and it is still widely used. Consider a 10-item anxiety questionnaire where each item is rated from 1 ("not at all") to 4 ("almost always"). CTT says: sum the 10 ratings into a total score ranging from 10 to 40, and treat that number as the person's anxiety level. A higher total means they are more anxious. Cronbach's alpha then tells you whether the items are consistent enough with each other that the sum is a trustworthy measurement.

But CTT has some awkward limitations. The properties it estimates for your items, like how hard the questions are to answer or how much they are related to the total sum score depend on the particular sample of people you tested.

Measure a hard item in a high-ability group and it looks easy compared to an average-ability group of people. So everyone who is tested is treated as equally precise. A person who scored at the very top and a person who scored right in the middle have the same standard error. And if you want to compare scores from two slightly different versions of a test, CTT makes that surprisingly hard. Put simply, CTT gives you one number describing your whole test, applied equally to everyone, estimated from whoever happened to be in your sample. This is a lot of assumptions baked in.

**Item response theory (IRT)** is a more modern approach that addresses these limitations directly. Instead of working with total scores like CTT, IRT models the probability that a person endorses each individual item as a function of where they stand on the latent trait. For example, rather than saying "this person scored 27 out of 40," IRT says something more like: "given where this person sits on the anxiety construct, there is an 85% chance they endorsed the item 'I feel nervous in social situations' and only a 20% chance they endorsed the item 'I cannot leave my house due to fear'."

The result is a framework that tells you not just how reliable a test is overall, but which items are doing the most work, where on the trait continuum the test is most precise, and why two people with the same total score might actually be quite different from each other.

This page covers the conceptual foundations of IRT: what the models are, what their parameters mean, and how to think about measurement precision along the trait continuum. No prior exposure to CTT is assumed, though readers who have encountered it will find natural points of connection throughout. The follow-up page, IRT in Practice, puts these ideas to work with real data and R code.

------------------------------------------------------------------------

## 1. From CTT to IRT: Why Make the Switch?

### 1.1 What CTT Gets Right (and Wrong)

Classical test theory starts from a deceptively simple idea: an observed score $X$ is the sum of a true score $T$ and an error term $E$. In other words, an individual's score on any given measurement is an imperfect reflection of the underlying thing you are trying to measure. There will be error associated with your measurement (its not possible to know what someone's *true* anxiety would ever be). This relationship can be formulated as

$$X = T + E$$

For more information, refer back to the Classical Test Theory lesson.

CTT is simple, robust, and produces actionable information with modest sample sizes. The problem with it is that it is limited in ways that become consequential when you start asking pointed questions.

### 1.2 Limitations of CTT

**Sample dependence.** In CTT, item and test statistics are estimated from a particular sample. Item difficulty (the proportion of correct responses) and item-test correlations will differ across samples with different trait distributions. Or in other words, a "hard" item in one group of people might look "moderate" if the second group are high in ability. This means you cannot directly compare item statistics across studies unless the samples are comparable.

**A constant standard error.** The standard error of measurement in CTT is the same for everyone, which means the test is treated as equally precise for everyone, regardless of where they actually fall on the trait. E.g., A highly anxious person and a mildly anxious person are assumed to have the same margin of error around their score. The variation in responses on an anxiety questionnaire should be higher in the region of the trait where items are most targeted, and lower where items are sparse or poorly matched to respondents. Think of a bathroom scale that is highly accurate between 100 and 200 pounds but wildly imprecise outside that range. CTT would report one average accuracy figure for that scale and apply it to everyone who stepped on it. That is exactly what a constant standard error of measurement does for a psychological test: it collapses real variation in measurement precision into a single number and pretends it applies equally everywhere.

**Test-specific scores.** A person's estimated true score is tied to the particular test used to estimate it. If you want to compare scores from two different versions of a measure, you need equating procedures that rely on assumptions CTT cannot always satisfy. If the items change even slightly, the score scale changes with them. In other words, if a clinic switches from one anxiety questionnaire to a slightly revised version midway through a treatment study, CTT gives you no principled way to know whether a patient's score change reflects genuine improvement or just the fact that the new items happen to run a little easier.

**No model to evaluate.** CTT does not make strong predictions about individual item responses. You cannot formally test whether the probability of endorsing an item behaves in a particular way as a function of the underlying trait. There is no model fit to evaluate. To make this tangible: IRT will tell you if an item is behaving strangely, for example, if people with high anxiety are no more likely to endorse "I feel nervous constantly" than people with low anxiety, which would suggest something is wrong with that item. CTT has no equivalent check. It summarizes how your items hang together as a set, but it never looks under the hood at whether each individual item is functioning the way a good measurement item should. So some items might be pulling more of the weight than others.

### 1.3 CTT vs. IRT: A Side-by-Side

The table below summarizes the key differences between the two frameworks. This table gives some technical distinctions between the two methods.

| Feature | CTT | IRT |
|------------------------|------------------------|------------------------|
| Item/test statistics | Sample-dependent | Invariant within linear transformation |
| Standard error of measurement | Constant for all persons | Varies across trait levels |
| Score scale | Test-specific | Common latent scale ($\theta$) |
| Reliability summary | Single coefficient | Information function |
| Test equating | Requires parallel tests | Links via shared $\theta$ scale |
| Norm vs. content referencing | Norm referencing only | Both possible |
| Missing data | Problematic | Handled via likelihood |
| Mixed item formats | Difficult | Straightforward |
| Model fit | Cannot be evaluated | Formally testable |

We will cover what each of these mean for the IRT models. The central IRT insight is that persons and items can be placed on the same scale that we call $\theta$.

In an anxiety questionnaire, an item like "I sometimes feel a little on edge before important events" is most useful for distinguishing people with mild anxiety from people with none at all. An item like "I cannot leave my house because of fear" only starts discriminating meaningfully among people who are already quite anxious. A person with moderate anxiety is being measured well by the first type of item and barely at all by the second. IRT puts both the items and the people on the same scale, so you can see exactly where your measurement is sharp and where it is blurry, which items are doing real work for your target population and which are dead weight, and whether the test as a whole is well matched to the people you actually care about measuring.

------------------------------------------------------------------------

## 2. The Item Response Curve

### 2.1 The Latent Trait $\theta$

The foundation of IRT is the latent trait $\theta$ (generically named as theta). This is the unobserved variable underlying item responses, which is the thing the test is designed to measure. In ability testing, $\theta$ is proficiency. In personality and psychopathology research, $\theta$ is the level of anxiety, depression, extraversion, and so on.

For identification purposes, $\theta$ is standardized to have mean zero and standard deviation one in the population: $\theta \sim N(0, 1)$. Think of it like a z-score on the latent trait (review Z scores if needed). A person with $\theta = 0$ is at the population average of that trait. A person with $\theta = 1.5$ is about 1.5 standard deviations above average. A person with $\theta = -2$ is well below average.

The key IRT assumption is that $\theta$ is the only thing driving item responses. Once you know a person's trait level, knowing anything else about them like their demographics, their responses to other items adds no additional information about how they will respond to any given item. We will discuss this in greater detail in the IRT in Practice page.

### 2.2 Parameters: Slope, Location, and Intercept

The remainder of this section focuses on **dichotomous items**, meaning items with exactly two possible responses: correct or incorrect, endorsed or not endorsed, yes or no. This covers the majority of ability testing items and many clinical screening tools. Psychological scales that use Likert-type response formats with three, four, five, or more ordered categories are called **polytomous items** and require an extension of the framework described here. The core logic of slope and location carries over completely to polytomous models, but we will cover polytomous items more in the next lesson. For now, building your intuition on the simpler dichotomous case will make everything that follows considerably easier to understand.

Every dichotomous IRT model describes the probability of endorsing an item as a function of $\theta$ and a small set of item parameters. The three most important are:

$\alpha$ (alpha): the discrimination parameter. This controls how sharply an item separates people who have a lot of the trait from people who do not (the slope of the item response curve, discussed later). A highly discriminating item is one where crossing the threshold makes a big difference: people just above it are very likely to endorse the item and people just below it are very unlikely to. A weakly discriminating item is more of a shrug: people across a wide range of trait levels all have a similar middling probability of endorsing it, which means it is not doing much to tell people apart. So the item scale can't tell the difference between a very anxious person and a moderately anxious person. In psychological research, typical values of $\alpha$ range from about 1.0 to 2.5, and the difference matters more than you might expect: two items targeting the same region of the anxiety continuum, one with $\alpha = 2.0$ and one with $\alpha = 1.0$, will not perform equally. The more discriminating item carries four times the measurement information at that region of the trait.

E.g., for a dichotomous anxiety item like "I cannot leave my house because of fear" with $\beta = 1.8$, a person needs to be about 1.8 standard deviations above the average anxiety level before they have a 50% chance of endorsing it. For most people in a general population sample, that item is essentially out of range.

$\beta$ (beta): the location or difficulty parameter. This is the point on the $\theta$ scale where a person has a 50% chance of endorsing the item. Recall your time in school and that some exam questions were clearly harder than others. This was usually intentional. The easy questions were meant to separate students who understood nothing from students with basic competence. The hard questions were meant to separate the B level students from the A level students. In IRT, $\beta$ captures exactly that: it tells you which region of the trait continuum an item is tuned to discriminate in. A high $\beta$ item is only useful for telling people apart near the top of the scale, just like a hard exam question only starts differentiating students once they already know most of the material. In ability testing $\beta$ is called the *difficulty* parameter for exactly this reason. In personality research it is called the *location* or *threshold* parameter because there is no right or wrong answer, just a point along the trait where endorsement becomes more likely than not.

$\alpha$ and $\beta$ are connected but distinct concepts. Imagine two anxiety items that both have $\beta = 1.5$, meaning both require a fairly high level of anxiety before endorsement becomes likely. The first item, "I cannot leave my house because of fear," has $\alpha = 2.2$, so it discriminates sharply. People just below $\theta = 1.5$ rarely endorse it and people just above endorse it readily. The second item, "I often feel that something bad is about to happen," also has $\beta = 1.5$ but $\alpha = 0.9$, so it is much more gradual. In this case, people across a wide range of anxiety levels have a similar moderate probability of endorsing it, which means it is doing relatively little to tell anxious people apart even though it is targeting the same region of the anxiety. We will see exactly what this looks like visually when we introduce the item response curve in Section 2.3.

$\gamma$ (gamma): the intercept. This is the expected log-odds (think of log-odds as a transformed probability: a value of 0 means the person has a 50% chance of endorsing the item, negative values mean less than 50%, and positive values mean greater than 50%) of endorsement when $\theta = 0$, meaning for a person sitting exactly at the population average on the trait. You can think of it as the item's baseline: before you know anything about how discriminating the item is or where it is located, $\gamma$ tells you the starting probability of endorsement for an average person. It is not an independent parameter though, it is completely determined by $\alpha$ and $\beta$ through the relationship $\gamma = -\alpha\beta$. So if an item has a high $\beta$ parameter (it is a hard or rarely endorsed item), $\gamma$ will be a large negative number, reflecting the fact that an average person has a low probability of endorsing it. If the item is easy or frequently endorsed, $\gamma$ will be closer to zero or even positive.

You will encounter $\gamma$ mostly when reading raw `mirt` output, where it appears as the column `d` in intercept parameterization. Most of the time you will want to convert to IRT parameterization using `IRTpars = TRUE`, which gives you the more interpretable $\alpha$ and $\beta$ directly. But it is worth knowing what `d` represents so the default output does not look like a mystery when you first see it.

### 2.3 Item Response Curves

An item response curve (IRC) is simply a plot of the probability of endorsing an item on the y-axis against trait level ($\theta$) on the x-axis. It answers the straightforward question of, for a person at any given level of the trait, what is the probability they endorse this item?

The shape of an IRC follows an S-curve. At the low end of the trait, the probability of endorsement is near zero. As trait level increases, the probability rises, slowly at first, then more steeply through a middle transition zone, then leveling off near one at the high end. For an anxiety item like "I cannot leave my house because of fear," the curve would be flat and low across most of the trait range and only start rising steeply for people with quite high anxiety. For an item like "I sometimes feel nervous," the curve would start rising much earlier, reflecting the fact that even moderately anxious people frequently endorse it.

Two features of the curve do most of the work. The location of the steep transition in the middle tells you where on the trait the item is most useful, which is $\beta$. The steepness of that transition tells you how sharply the item discriminates between people just below and just above that point, which is $\alpha$. The plot below shows a single item response curve to make these features concrete before we move into the models.

Below is a fake item response curve that serves as a demonstration. The parameters used here are illustrative: the location $\beta = 0.21$ is approximated from the empirical endorsement rate of LSAT Item 3, and the discrimination $\alpha = 1.5$ is a plausible but arbitrary value chosen to produce a reasonable curve shape. Neither is a formally estimated IRT parameter. We will replace them with properly estimated values in the IRT in Practice page.

```{r irc-intro, echo = F}
p_item  <- 0.533                     # empirical endorsement rate of LSAT item 3
beta    <- qlogis(p_item)            # rough beta estimate (= 0.21)
alpha   <- 1.5                       # arbitrary alpha for illustration

theta   <- seq(-4, 4, length.out = 200)
p       <- exp(alpha * (theta - beta)) / (1 + exp(alpha * (theta - beta)))
irc_df  <- data.frame(theta = theta, prob = p)

ggplot(irc_df, aes(x = theta, y = prob)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_vline(xintercept = beta, linetype = "dashed", color = "firebrick") +
  geom_hline(yintercept = 0.5,  linetype = "dashed", color = "firebrick") +
  annotate("text", x = beta + 0.15, y = 0.1,
           label = paste0("β = ", round(beta, 2)), color = "firebrick", hjust = 0) +
  labs(
    x        = expression(theta),
    y        = "P(X = 1 | θ)",
    title    = "A Single Item Response Curve",
    subtitle = "Dashed lines mark β, where P(endorsement) = .50"
  ) +
  theme_minimal()
```

The dashed lines mark $\beta = 0.21$, which is the point where the curve crosses the 50% probability line. Think of it as the item's tipping point: people to the left of that vertical line are more likely than not to get the item wrong, and people to the right are more likely than not to get it right. Since $\beta = 0.21$ is close to $\theta = 0$, this is a roughly average difficulty item. A person of average ability (recall that, since $\theta$ is standardized, $\theta = 0$ means the person has average competency) has about 50:50 chance of getting it correct. Notice also how the curve rises steeply right around the $\beta$ tipping point and flattens out toward the extremes. That steepness is $\alpha$ saying that the item is most useful for telling people apart right around the middle of the trait range, and much less useful for people who are very high or very low on the trait.

Everything that follows in this section is really just a description of how different IRT models handle the shape and positioning of this curve across a set of items.

### 2.4 The One-Parameter Logistic Model (1PL / Rasch)

The simplest IRT model constrains all discrimination parameters to the same value, usually 1. So items differ only in their location. This is often called a **1PL model** (short for *one-parameter logistic model*). Conceptually, this is like saying that every question on an exam is equally good at discriminating between students, and the only thing that varies is difficulty. Some questions are hard and some are easy, but all of them are assumed to do the same quality of job separating students who know the material from those who do not. In practice this is a strong assumption, and one that real data often violate.

Mathematically, the curve is described as

$$
P(X = 1 \mid \theta) = \frac{\exp(\theta - \beta)}{1 + \exp(\theta - \beta)}
$$

The only thing that varies across items is $\beta$, the location. As a result, all item response curves under the 1PL have exactly the same shape and steepness, just shifted left or right along the $\theta$ axis depending on how easy or hard the item is. An easy item has a curve shifted to the left, meaning even people with low trait levels have a reasonable chance of endorsing it. A hard item has a curve shifted to the right, meaning you need to be well above average on the trait before endorsement becomes likely.

Because all slopes are equal, item response curves for different items are parallel. They have the same shape and steepness, just shifted left or right along the $\theta$ axis. This can be understood better as item information, which is discussed in Section 3.

The 1PL (and its close relative the **Rasch model**) has appealing mathematical properties. Person and item parameters are separable, which in plain terms means the difficulty of an item does not depend on who you happened to test, and the estimate of a person's ability does not depend on which particular items they happened to receive. The raw total score is also a sufficient statistic for $\theta$, meaning that under the 1PL your score is entirely determined by how many items you got right, not which items you got right. Get items 1, 3, and 5 correct or items 2, 4, and 5 correct and you will get the same $\theta$ estimate either way.

In practice, the equal-discrimination assumption is quite strong and often fails empirically. On an anxiety scale, an item like "I cannot leave my house because of fear" might discriminate very sharply at the high end of the trait, while an item like "I sometimes feel a little stressed" might be much more gradual and diffuse. Treating those two items as equally discriminating is a meaningful constraint that real data frequently reject. The 1PL is a useful baseline, but you should expect the 2PL to fit most real psychological data better, discussed next.

Below is a fake example IRC for three difficulty parameters for an "easy" item, a "moderate" item, and a "hard" item.

```{r irc-1pl, echo = F}
theta <- seq(-4, 4, length.out = 200)
beta_vals <- c(-1, 0, 1.5)
labels <- c("Easy (β = -1)", "Moderate (β = 0)", "Hard (β = 1.5)")

irc_1pl <- do.call(rbind, lapply(seq_along(beta_vals), function(i) {
  p <- exp(theta - beta_vals[i]) / (1 + exp(theta - beta_vals[i]))
  data.frame(theta = theta, prob = p, item = labels[i])
}))

ggplot(irc_1pl, aes(x = theta, y = prob, color = item)) +
  geom_line(linewidth = 1) +
  labs(x = expression(theta), y = "P(X = 1 | θ)",
       title = "1PL Item Response Curves", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Notice that the three curves are parallel with same shape and steepness. They are just shifted horizontally. That is the defining feature of the 1PL. Items are distinguished only by where they sit on the trait axis, not by how sharply they discriminate.

### 2.5 The Two-Parameter Logistic Model (2PL)

The 2PL frees the discrimination parameter to vary across items. Items now have both a slope and a location. Mathematically, the curve is described as

$$
P(X = 1 \mid \theta) = \frac{\exp(\alpha(\theta - \beta))}{1 + \exp(\alpha(\theta - \beta))}
$$

Unlike the 1PL, now the $\alpha$ term enters the function and allows each curve to differ in their slope and discrimination ability. A highly discriminating item is preferred over a less discriminating item for people near the item's location ($\beta$), but provides less information for people far from it.

An example of a fake 2PL plot is given below for a "high", a "moderate", and a "low" discriminatory item.

```{r irc-2pl, echo = F}
alpha_vals <- c(0.5, 1.0, 2.0)
beta_fixed <- 0
labels_a <- c("Low discrimination (α = 0.5)",
               "Moderate discrimination (α = 1.0)",
               "High discrimination (α = 2.0)")

irc_2pl <- do.call(rbind, lapply(seq_along(alpha_vals), function(i) {
  p <- exp(alpha_vals[i] * (theta - beta_fixed)) /
       (1 + exp(alpha_vals[i] * (theta - beta_fixed)))
  data.frame(theta = theta, prob = p, item = labels_a[i])
}))

ggplot(irc_2pl, aes(x = theta, y = prob, color = item)) +
  geom_line(linewidth = 1) +
  labs(x = expression(theta), y = "P(X = 1 | θ)",
       title = "2PL Item Response Curves: Effect of Discrimination", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

All three curves cross the .50 probability line at $\theta = 0$. The three items share the same location ($\beta = 0$), but notice how much steeper the high-discrimination curve is ($\alpha = 2$). A highly discriminating item draws a sharp line between people just below and just above the threshold. A low-discrimination item is gradual. It provides some information across a wide range of $\theta$, but very little at any particular point.

Here is how these two might in a questionnaire. An item with high $\alpha = 2$ and low $\beta = .2$ has a steep curve centered in the low anxiety region, meaning it is most precise for people near the bottom of the trait. If you are building a test and you want the sharpest possible measurement near the boundary between no anxiety and mild anxiety, you would want an item with these values. Similarly, if you want an item that is good at discriminating high anxiety from debilitating anxiety, you would need an item like $\alpha = 2$ and $\beta = .9$.

It is much rarer when you would want items with low $\alpha$ discrimination. Some examples, however, might include situations where you are building a very short scale intended to give a rough sense of where someone falls across the entire trait range, rather than making precise distinctions at any particular point. A single broad item that provides modest information everywhere is sometimes more useful than a sharp item that is highly informative in one narrow band and nearly useless everywhere else. Low discrimination items also tend to appear in measures of fuzzy or multifaceted constructs like quality of life or general wellbeing, where the construct itself resists sharp boundaries and a gradual, diffuse item may actually be a more honest reflection of how the trait works in practice.

### 2.6 The 3PL and 4PL Models

In achievement testing, guessing is a real concern. Think about a multiple choice exam with four answer options. A student who has no idea what the correct answer is still has a 25% chance of getting it right just by picking randomly. This means that even for people at the very bottom of the ability distribution, the probability of a correct response never quite reaches zero. That nonzero floor is called the lower asymptote, and the standard 2PL has no way to represent it. If you look at the plot above, the 2PL curve just keeps dropping toward zero as ability decreases, which is unrealistic when blind guessing is possible.

The 3PL model handles this with a new "guessing" parameter called $c$. This is put into the probability curve in the following manner

$$
P(X = 1 \mid \theta) = c + (1 - c) \cdot \frac{\exp(\alpha(\theta - \beta))}{1 + \exp(\alpha(\theta - \beta))}
$$

The parameter $c$ describes the lower asymptote "guess" probability. For a four-option multiple-choice item, $c$ would plausibly be .25 since an individual would only have a $1/4$ chance of getting any answer correctly by just guessing.

The 4PL model adds an upper asymptote parameter $d$ for **inattention**. Think about a student who knows the material cold but misreads the question, runs out of time, or simply clicks the wrong answer by accident. For most items this is a minor concern, but for very easy items where nearly everyone should succeed, even a small rate of careless errors means the curve never quite reaches 1.0 at the top of the trait range. The 4PL adds an upper asymptote parameter to the probability curve function $d$ to capture this reality:

$$
P(X = 1 \mid \theta) = c + (1 - d) \cdot \frac{\exp(\alpha(\theta - \beta))}{1 + \exp(\alpha(\theta - \beta))}
$$

Both $c$ and $d$ are on the probability metric (bounded between 0 and 1). The 3PL is common in educational measurement but less relevant for personality and clinical research where guessing is minimal. For most psychological applications, the 2PL is the appropriate model.

Rarely would a psychologist introduce $d$, but one scenario where it becomes relevant is high-stakes clinical assessment, where a patient who is genuinely at the top of the anxiety distribution might still occasionally fail to endorse an item simply because the wording did not resonate with how they personally experience the symptom. Sometimes, it is important to keep potential false-negatives into account.

One final note before moving on. To rehash, everything in this section has assumed items with only two possible responses, endorsed or not endorsed (or correct vs not correct). The models covered in the next lesson extend these same ideas to items with multiple ordered response categories, where each boundary between adjacent categories gets its own $\beta$ location parameter while $\alpha$ remains shared across all of them. The core logic of slope and location carries over completely, there is just more things in one item to endorse or not.

------------------------------------------------------------------------

## 3. Item and Test Information

### 3.1 Item Information Functions

One of the most useful features of IRT is that it tells you exactly how much information about $\theta$ each item provides and *where* along the trait continuum it provides that information. Think of information as measurement precision: a high information value at a particular $\theta$ means the item is doing a good job of pinning down exactly where on the trait a person sits. A low information value means the item is contributing little, essentially shrugging at that region of the scale. In practice this matters enormously. A clinical depression screener administered to a general population sample might have most of its items clustered in the moderate range of the trait, providing excellent precision for distinguishing mildly from moderately depressed people but doing almost nothing to differentiate someone with severe depression from someone who is completely debilitated. IRT makes that invisible inefficiency visible by showing you exactly where along the trait your test is sharp and where it is essentially guessing.

Information helps in this situation by turning a vague sense that your test might not work well at the extremes into a precise, actionable picture of exactly where it falls short and what you would need to fix it.

First I will go into some technical detail but try to bring this back to actionable advice. The item information function $I_i(\theta)$ quantifies this:

-   **1PL:** $I_i(\theta) = P_i(\theta) \cdot Q_i(\theta)$
-   **2PL:** $I_i(\theta) = \alpha_i^2 \cdot P_i(\theta) \cdot Q_i(\theta)$

In both cases $P_i(\theta)$ is the probability of endorsing the item at a given trait level and $Q_i(\theta) = 1 - P_i(\theta)$ is the probability of not endorsing it. Their product $P \cdot Q$ is highest when $P = .50$, meaning an item is most informative right at the point where endorsement becomes a coin flip. Move too far above or below the item's location and one of those probabilities approaches zero, dragging the product down with it. The 2PL adds $\alpha^2$ as a multiplier, which is why discrimination matters so much: an item with $\alpha = 2.0$ has $\alpha^2 = 4$, meaning it carries four times the information of an item with $\alpha = 1.0$ at the same location, even though both items have the same $P \cdot Q$ at their peak. Discrimination does not change where the item is most informative, it just determines how much information it delivers there.

Information is always highest near an item's location parameter and falls off on both sides. This makes sense as the item was built with that location in mind. A highly discriminating item produces a tall, narrow information spike, and a weakly discriminating item produces a low, broad hill.

An example of a fake information curve for three different items is given below for demonstration.

```{r item-info, echo = F}
items <- list(
  list(alpha = 1.5, beta = -1,   label = "α=1.5, β=-1"),
  list(alpha = 2.0, beta =  0,   label = "α=2.0, β=0"),
  list(alpha = 0.8, beta =  1.5, label = "α=0.8, β=1.5")
)

info_data <- do.call(rbind, lapply(items, function(it) {
  p  <- exp(it$alpha * (theta - it$beta)) /
        (1 + exp(it$alpha * (theta - it$beta)))
  ii <- it$alpha^2 * p * (1 - p)
  data.frame(theta = theta, info = ii, item = it$label)
}))

ggplot(info_data, aes(x = theta, y = info, color = item)) +
  geom_line(linewidth = 1) +
  labs(x = expression(theta), y = "Item Information",
       title = "Item Information Functions", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The item with $\alpha = 2.0$ has the highest peak, nearly four times the height of a slope-1.0 item at the same location. The item with $\alpha = 0.8$ is spread out and low, indicating weak discrimination across the whole trait range.

The intuition behind the information formula connects directly back to trait discrimination, and understanding it has real practical value for anyone designing or evaluating a psychological measure.

Consider two people responding to the item "I have panic attacks every day." One has moderate anxiety, the other has severe anxiety. If this is a highly discriminating item, the moderately anxious person almost certainly says no and the severely anxious person almost certainly says yes. That clean separation is what makes the response informative: a clinician who knows how a patient answered that item has learned something meaningful about where they sit on the anxiety continuum. Now imagine a poorly worded version of the same item where both groups have about a 50/50 chance of endorsing it regardless of their actual severity. The response still comes in, but it tells you almost nothing. You administered an item, used up a patient's time and attention (not to mention your precious grant budget), and walked away knowing barely more than you did before.

This is where item information becomes a practical tool rather than just a statistical concept. Suppose you are a psychologist working in a hospital and you are evaluating whether your current depression screener is fit for purpose. The information function might reveal that your 20-item scale delivers excellent precision for patients in the mild to moderate range but essentially loses its ability to differentiate among patients at the severe end, which is exactly where your clinical decisions are most consequential. That tells you something actionable: you do not need more items, you need better-targeted items. Specifically, you need items with high discrimination and high location parameters, items that only start to differentiate once a patient is already quite severely depressed. Without the information function, you would never know this from a reliability coefficient alone. Alpha might be .90 and you would feel confident in your scale, all while missing the patients who need the most precise assessment.

### 3.2 Test Information and the Standard Error of Measurement

Once you have item information functions for each item in your scale, combining them can give you the overall information of your test. I.e., how well your test informs you about the latent psychological trait. To do this is straightforward: you simply add them up. The resulting **test information function** tells you the total measurement precision your scale delivers at every point along the trait continuum. Think of it like stacking the contributions of each item on top of one another. An item that is highly informative near $\theta = 0$ adds a tall peak to that region of the test information curve. An item targeted at $\theta = 2$ adds its contribution further up the scale. The final curve reflects the collective measurement capacity of the whole instrument, and its shape tells you immediately where your test is doing its best work and where it is running thin.

The **standard error of measurement** follows directly from test information and is its practical translation. High test information at a given $\theta$ means low standard error, meaning your score estimates of the latent trait are precise in that region. Low test information means high standard error, meaning your estimates of the latent trait are uncertain. This is the IRT answer to the CTT limitation we discussed in Section 1: rather than a single SEM that gets applied uniformly to everyone, you now have a standard error that varies across the trait and honestly reflects the fact that your test measures some people more precisely than others.

Mathematically, test information is the sum of item information functions:

$$TINFO(\theta) = \sum_{i=1}^{n} I_i(\theta)$$

And the standard error of measurement at any given trait level is:

$$SE(\hat{\theta}) = \frac{1}{\sqrt{TINFO(\theta)}}$$

The standard error of measurement varies across $\theta$. A test with items clustered around $\beta = 0$ is highly precise for people near the population mean but much less precise for people with extreme trait values. A clinical screening tool might deliberately place items at higher $\beta$ values to be most precise in the range where diagnostic decisions are made. It all depends on the purpose of your test and what population you are targeting.

An example plot for test information and standard error of measurement is given below for a fake test.

```{r test-info, echo = F}
tinfo <- aggregate(info ~ theta, data = info_data, FUN = sum)
tinfo$se <- 1 / sqrt(tinfo$info)

p_info <- ggplot(tinfo, aes(x = theta, y = info)) +
  geom_line(linewidth = 1, color = "steelblue") +
  labs(x = expression(theta), y = "Test Information",
       title = "Test Information Function") +
  theme_minimal()

p_se <- ggplot(tinfo, aes(x = theta, y = se)) +
  geom_line(linewidth = 1, color = "firebrick") +
  labs(x = expression(theta), y = "Standard Error",
       title = "Standard Error of Measurement") +
  theme_minimal()

print(p_info)
print(p_se)
```

As one can see, the standard error of measurement curve is the mirror image of the information curve. Where we can see information in the test is high, the standard error is low. And where the test information is low, there is much more error. This can inform us that the test measures most precisely where it has the most items with high discrimination located.

In clinical psychology, this has direct implications for how you design and evaluate assessment tools. A depression screener intended for use in a primary care setting, where the goal is to catch patients who might be slipping into clinical range, should have its information peak somewhere in the mild to moderate region of the trait. If the test information curve shows a peak at $\theta = 0$ but drops off sharply above $\theta = 1.5$, a clinician using it to identify severely depressed patients is essentially working with a blunt instrument when they need a sharp knife. The standard error curve will show this honestly. Large standard errors at the high end of the trait mean that two patients who are genuinely quite different in their depression severity (one who struggles to get out of bed most days and one who is completely unable to function) might receive scores so close together that the test cannot distinguish between them. For a clinician trying to make decisions about level of care or treatment intensity, that is a meaningful failure.

### 3.3 Marginal Reliability

The concepts of reliability in item response theory (IRT) and classical test theory (CTT) are more connected than they might initially appear. Recall from Section 1 that CTT reliability is a single number summarizing how much of the observed score variance is attributable to true score variance. That number is useful but it collapses everything into one figure and applies it uniformly to everyone. Marginal reliability is the IRT equivalent: it asks what the average reliability of the test is across the full distribution of $\theta$ in the population, integrating the information function over where people actually tend to fall on the trait. This is mathematically given as

$$\bar{\rho} = 1 - \frac{1}{\overline{TINFO}}$$

Where $\overline{TINFO}$ is the average test information across the population distribution of $\theta$. The formula has an intuitive structure: when average test information is high, the fraction $\frac{1}{\overline{TINFO}}$ is small, and marginal reliability approaches 1. When test information is low on average, that fraction grows and reliability drops. In this sense marginal reliability is doing exactly what CTT reliability does, summarizing overall precision as a single number between 0 and 1, but it is built from the information function rather than from observed score variances.

The key difference is that marginal reliability is sensitive to the match between your test and your population. To make this concrete: imagine a research team develops a social anxiety scale by piloting it on a clinical sample of patients seeking treatment for anxiety disorders. Most of the participants in that pilot sample have moderate to severe social anxiety, so the items end up being calibrated toward the higher end of the trait. The scale has good marginal reliability in that clinical sample, perhaps around .90, and the team publishes it feeling confident.

A few years later, a different research group wants to use the same scale in a large community survey where most participants have little to no social anxiety. When they compute marginal reliability in their sample, it is noticeably lower, perhaps around .72, even though they are using the exact same items. The CTT reliability coefficient would tell them something went wrong but gives no indication of why. The test information function tells them immediately: the items are delivering their precision up at $\theta = 1.5$ and above, but the bulk of their community sample is sitting near $\theta = -0.5$, a region where the scale is barely measuring anything. The scale was not poorly designed, it was just designed for a different population than the one they are using it with.

The researchers would get to this conclusion by plotting the test information curve alongside the distribution of $\theta$ estimates from their sample. The information curve shows where the test is precise. The $\theta$ distribution shows where their participants actually are. When you overlay the two, the mismatch becomes immediately obvious: a tall information peak sitting at $\theta = 1.5$ with most of the sample clustered near $\theta = -0.5$ tells the whole story at a glance. The region where the test is working hardest is almost entirely unpopulated, and the region where most participants sit is where the information curve has already flattened out near zero.

The bottom line is this: you cannot simply take a test developed for one context and apply it to another. The population you are measuring matters just as much as the items themselves. We will see how to produce and interpret these plots with real data in the IRT in Practice page.

------------------------------------------------------------------------

## Well Done!

You have completed the IRT Foundations page. Here is a summary of what was covered:

-   The core limitations of classical test theory: sample dependence of item statistics, a constant standard error of measurement applied to everyone, test-specific scores that cannot be easily compared across versions, and the absence of any formal model to evaluate
-   The CTT vs. IRT comparison table and the central IRT insight that persons and items can be placed on the same $\theta$ scale
-   The latent trait $\theta$ as a standardized continuous variable with mean 0 and standard deviation 1 in the population
-   The three key item parameters: $\alpha$ (discrimination, how sharply the item separates people at its threshold), $\beta$ (location, where on the trait the item is most useful), and $\gamma$ (the intercept, which is fully determined by $\alpha$ and $\beta$)
-   The item response curve (IRC) as the fundamental building block of IRT: an S-shaped curve describing the probability of endorsement as a function of $\theta$
-   The one-parameter logistic model (1PL), where all items share the same slope and differ only in location, producing parallel item response curves
-   The two-parameter logistic model (2PL), where items differ in both slope and location, allowing curves to cross and capturing real variation in item quality
-   The 3PL and 4PL extensions for guessing and inattention, most relevant in achievement testing contexts
-   Item information functions and how discrimination ($\alpha^2$) and location ($\beta$) jointly determine where and how much information an item provides
-   Test information as the sum of item information functions, and the standard error of measurement as its inverse square root, which varies across $\theta$ rather than being constant
-   Marginal reliability as the IRT analog of CTT reliability, and why it is sensitive to the match between test design and the population being measured

The next page, IRT in Practice, puts these concepts to work with real data and R code, starting with assumption checking and moving through model fitting, polytomous models, computerized adaptive testing, and differential item functioning.
