---
title: "Item Response Theory"
author: "Michael Woller"
draft: true
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(mirt)
library(mokken)
library(ltm)
library(psych)
library(psychTools)
library(ggplot2)

# Dichotomous IRT examples: Law School Admissions Test (ltm package)
# 5 items, N = 1000, dichotomous scoring (0 = incorrect, 1 = correct)
data(LSAT, package = "ltm")

# Polytomous GRM examples: Big Five Inventory extraversion items (psychTools package)
# Selecting the 5 extraversion items (E1-E5) from the bfi dataset
# Items are rated 1-6 (Very Inaccurate to Very Accurate), N = 2800
data(bfi, package = "psychTools")
bfi_extra <- bfi[, c("E1", "E2", "E3", "E4", "E5")]
bfi_extra <- na.omit(bfi_extra)

# Polytomous NRM examples: Science attitudes dataset (mirt package)
# 9 polytomous items rated on a 4-point scale, N = 392
data(Science, package = "mirt")
```

## Introduction

Imagine you are trying to measure something you cannot directly observe, like how anxious someone is, how much they agree with a political ideology, or how strong their reading ability is. You cannot just look at a person and read off their anxiety level the way you might read a thermometer. Instead, you ask them a set of questions, and you use their answers as a window into the underlying thing you actually care about. This is the core of psychological measurement as we've discussed before.

As we discussed previously, for most of the twentieth century, the dominant answer came from **classical test theory (CTT)**. CTT is intuitive: add up a person's item scores (questions on a test that hav ea point value), treat the total as their measurement, and use statistics like Cronbach's alpha to tell you how much random noise is contaminating the signal. It works, and it is still widely used. Consider a 10-item anxiety questionnaire where each item is rated from 1 ("not at all") to 4 ("almost always"). CTT says: sum the 10 ratings into a total score ranging from 10 to 40, and treat that number as the person's anxiety level. A higher total means they are more anxious. Cronbach's alpha then tells you whether the items are consistent enough with each other that the sum is a trustworthy measurement.

But CTT has some awkward limitations. The properties it estimates for your items, like how hard the questions are to answer or how much they are related to the total sum score depend on the particular sample of people you tested.

Measure a hard item in a high-ability group and it looks easy compared to a average-ability group of people. So everyone who is tested is treated as equally precise. A person who scored at the very top and a person who scored right in the middle have the same standard error. And if you want to compare scores from two slightly different versions of a test, CTT makes that surprisingly hard. Put simply, CTT gives you one number describing your whole test, applied equally to everyone, estimated from whoever happened to be in your sample. This is a lot of assumptions baked in.

**Item response theory (IRT)** is a more modern approach that addresses these limitations directly. Instead of working with total scores like CTT, IRT models the probability that a person endorses each individual item as a function of where they stand on the latent trait. For example, rather than saying "this person scored 27 out of 40," IRT says something more like: "given where this person sits on the anxiety construct, there is an 85% chance they endorsed the item 'I feel nervous in social situations' and only a 20% chance they endorsed the item 'I cannot leave my house due to fear'."

The result is a framework that tells you not just how reliable a test is overall, but which items are doing the most work, where on the trait continuum the test is most precise, and why two people with the same total score might actually be quite different from each other.

This page covers IRT from the ground up, working through the models, assumptions, and applied workflow for both dichotomous and polytomous data. The final two sections cover two of the most important applied extensions of IRT: computerized adaptive testing (CAT) and differential item functioning (DIF).

**Data used in this tutorial.** The dichotomous IRT examples use the `LSAT` dataset from the `ltm` package^1^. This is a classic IRT teaching dataset that contains responses from 1,000 respondents to five items from Section 6 of the Law School Admissions Test. All items are dichotomously scored (0 = incorrect, 1 = correct). The items vary in difficulty, with endorsement rates ranging from about .56 to .93, meaning some items are considerably easier than others (**endorsement rate** just means the percent of individuals who answered the item correctly).

The polytomous examples draw on two additional datasets. The first is the Big Five Inventory (BFI) from the `psychTools` package^2^, specifically the five extraversion items (`E1` through `E5`) rated on a 6-point scale from *very inaccurate* to *very accurate* (N=2,800 after removing missing data). Extraversion is the tendency toward sociability, assertiveness, and positive affect. The second dataset is named `Science` from the `mirt` package, which contains responses from 392 individuals to nine items measuring attitudes toward science and technology, each rated on a 4-point scale.

------------------------------------------------------------------------

## 1. From CTT to IRT: Why Make the Switch?

### 1.1 What CTT Gets Right (and Wrong)

Classical test theory starts from a deceptively simple idea: an observed score $X$ is the sum of a true score $T$ and an error term $E$. In other words, an individual's score on any given measurement is an imperfect reflection of the underlying thing you are trying to measure. There will be error associated with your measurement (its not possible to know what someone's *true* anxiety would ever be). This relatinship can be formulated as

$$X = T + E$$

For more information, refer back to the Classical Test Theory lesson.

CTT is simple, robust, and produces actionable information with modest sample sizes. The problem with it is that it is limited in ways that become consequential when you start asking pointed questions.

### 1.2 Limitations of CTT

**Sample dependence.** In CTT, item and test statistics are estimated from a particular sample. Item difficulty (the proportion of correct responses) and item-test correlations will differ across samples with different trait distributions. Or in other words, a "hard" item in one group of people might look "moderate" if the second group are high in ability. This means you cannot directly compare item statistics across studies unless the samples are comparable.

**A constant standard error.** The standard error of measurement in CTT is the same for everyone, which means the test is treated as equally precise for everyone, regardless of where they actually fall on the trait. E.g., A highly anxious person and a mildly anxious person are assumed to have the same margin of error around their score. The variation in responses on an anxiety questionnaire should be higher in the region of the trait where items are most targeted, and lower where items are sparse or poorly matched to respondents. Think of a bathroom scale that is highly accurate between 100 and 200 pounds but wildly imprecise outside that range. CTT would report one average accuracy figure for that scale and apply it to everyone who stepped on it. That is exactly what a constant standard error of measurement does for a psychological test: it collapses real variation in measurement precision into a single number and pretends it applies equally everywhere.

**Test-specific scores.** A person's estimated true score is tied to the particular test used to estimate it. If you want to compare scores from two different versions of a measure, you need equating procedures that rely on assumptions CTT cannot always satisfy. If the items change even slightly, the score scale changes with them. In other words, if a clinic switches from one anxiety questionnaire to a slightly revised version midway through a treatment study, CTT gives you no principled way to know whether a patient's score change reflects genuine improvement or just the fact that the new items happen to run a little easier.

**No model to evaluate.** CTT does not make strong predictions about individual item responses. You cannot formally test whether the probability of endorsing an item behaves in a particular way as a function of the underlying trait. There is no model fit to evaluate. To make this tangible: IRT will tell you if an item is behaving strangely, for example, if people with high anxiety are no more likely to endorse "I feel nervous constantly" than people with low anxiety, which would suggest something is wrong with that item. CTT has no equivalent check. It summarizes how your items hang together as a set, but it never looks under the hood at whether each individual item is functioning the way a good measurement item should. So some items might be pulling more of the weight than others.

### 1.3 CTT vs. IRT: A Side-by-Side

The table below summarizes the key differences between the two frameworks. This table gives some technical distinctions between the two methods.

| Feature | CTT | IRT |
|------------------------|------------------------|------------------------|
| Item/test statistics | Sample-dependent | Invariant within linear transformation |
| Standard error of measurement | Constant for all persons | Varies across trait levels |
| Score scale | Test-specific | Common latent scale ($\theta$) |
| Reliability summary | Single coefficient | Information function |
| Test equating | Requires parallel tests | Links via shared $\theta$ scale |
| Norm vs. content referencing | Norm referencing only | Both possible |
| Missing data | Problematic | Handled via likelihood |
| Mixed item formats | Difficult | Straightforward |
| Model fit | Cannot be evaluated | Formally testable |

We will cover what each of these mean for the IRT models. The central IRT insight is that persons and items can be placed on the same scale that we call $\theta$.

In an anxiety questionnaire, an item like "I sometimes feel a little on edge before important events" is most useful for distinguishing people with mild anxiety from people with none at all. An item like "I cannot leave my house because of fear" only starts discriminating meaningfully among people who are already quite anxious. A person with moderate anxiety is being measured well by the first type of item and barely at all by the second. IRT puts both the items and the people on the same scale, so you can see exactly where your measurement is sharp and where it is blurry, which items are doing real work for your target population and which are dead weight, and whether the test as a whole is well matched to the people you actually care about measuring.

------------------------------------------------------------------------

## 2. The Item Response Curve

### 2.1 The Latent Trait $\theta$

The foundation of IRT is the latent trait $\theta$ (generically named as theta). This is the unobserved variable underlying item responses, which is the thing the test is designed to measure. In ability testing, $\theta$ is proficiency. In personality and psychopathology research, $\theta$ is the level of anxiety, depression, extraversion, and so on.

For identification purposes, $\theta$ is standardized to have mean zero and standard deviation one in the population: $\theta \sim N(0, 1)$. Think of it like a z-score on the latent trait (review Z scores if needed). A person with $\theta = 0$ is at the population average of that trait. A person with $\theta = 1.5$ is about 1.5 standard deviations above average. A person with $\theta = -2$ is well below average.

The key IRT assumption is that $\theta$ is the only thing driving item responses. Once you know a person's trait level, knowing anything else about them like their demographics, their responses to other items adds no additional information about how they will respond to any given item. We will discuss this in greater detail in section 4.

### 2.2 Parameters: Slope, Location, and Intercept

Every IRT model describes the probability of endorsing an item as a function of $\theta$ and a small set of item parameters. The three most important are:

$\alpha$ (alpha): the discrimination parameter. This controls how sharply an item separates people who have a lot of the trait from people who do not (the slope of the item response curve, discussed later). A highly discriminating item is one where crossing the threshold makes a big difference: people just above it are very likely to endorse the item and people just below it are very unlikely to. A weakly discriminating item is more of a shrug: people across a wide range of trait levels all have a similar middling probability of endorsing it, which means it is not doing much to tell people apart. So the item scale can't tell the difference between a very anxious person and a moderately anxious person. In psychological research, typical values of $\alpha$ range from about 1.0 to 2.5, and the difference matters more than you might expect: two items targeting the same region of the anxiety continuum, one with $\alpha = 2.0$ and one with $\alpha = 1.0$, will not perform equally. The more discriminating item carries four times the measurement information at that region of the trait.

E.g., if the response of 3 (from 0-4) to the anxiety item "I cannot leave my house because of fear" has a $\beta$ of 1.8, that means a person needs to be about 1.8 standard deviations above the average anxiety level before they have a 50% chance of selecting that response or higher. That is a bit technical, but practically speaking, it just means that for most people in a general population sample, that item is essentially out of range. You have to be decently high in anxiety to start having an equal chance of putting a 3 or 2 down.

$\beta$ (beta): the location or difficulty parameter. This is the point on the $\theta$ scale where a person has a 50% chance of endorsing the item. Recall your time in school and that some exam questions were clearly harder than others. This was usually intentional. The easy questions were meant to separate students who understood nothing from students with basic competence. The hard questions were meant to separate the B level students from the A level students. In IRT, $\beta$ captures exactly that: it tells you which region of the trait continuum an item is tuned to discriminate in. A high $\beta$ item is only useful for telling people apart near the top of the scale, just like a hard exam question only starts differentiating students once they already know most of the material. In ability testing $\beta$ is called the *difficulty* parameter for exactly this reason. In personality research it is called the *location* or *threshold* parameter because there is no right or wrong answer, just a point along the trait where endorsement becomes more likely than not.

$\alpha$ and $\beta$ are connected but distinct concepts. Imagine two anxiety items that both have $\beta = 1.5$, meaning both require a fairly high level of anxiety before endorsement becomes likely. The first item, "I cannot leave my house because of fear," has $\alpha = 2.2$, so it discriminates sharply. People just below $\theta = 1.5$ rarely endorse it and people just above endorse it readily. The second item, "I often feel that something bad is about to happen," also has $\beta = 1.5$ but $\alpha = 0.9$, so it is much more gradual. In this case, people across a wide range of anxiety levels have a similar moderate probability of endorsing it, which means it is doing relatively little to tell anxious people apart even though it is targeting the same region of the anxiety. We will see exactly what this looks like visually when we introduce the item response curve in Section 2.

$\gamma$ (gamma): the intercept. This is the expected log-odds (think of log-odds as a transformed probability: a value of 0 means the person has a 50% chance of endorsing the item, negative values mean less than 50%, and positive values mean greater than 50%) of endorsement when $\theta = 0$, meaning for a person sitting exactly at the population average on the trait. You can think of it as the item's baseline: before you know anything about how discriminating the item is or where it is located, $\gamma$ tells you the starting probability of endorsement for an average person. It is not an independent parameter though, it is completely determined by $\alpha$ and $\beta$ through the relationship $\gamma = -\alpha\beta$. So if an item has a high $\beta$ parameter (it is a hard or rarely endorsed item), $\gamma$ will be a large negative number, reflecting the fact that an average person has a low probability of endorsing it. If the item is easy or frequently endorsed, $\gamma$ will be closer to zero or even positive.

### 2.3 Item Response Curves

An item response curve (IRC) is simply a plot of the probability of endorsing an item on the y-axis against trait level ($\theta$) on the x-axis. It answers the straightforward question of, for a person at any given level of the trait, what is the probability they endorse this item?

The shape of an IRC follows an S-curve. At the low end of the trait, the probability of endorsement is near zero. As trait level increases, the probability rises, slowly at first, then more steeply through a middle transition zone, then leveling off near one at the high end. For an anxiety item like "I cannot leave my house because of fear," the curve would be flat and low across most of the trait range and only start rising steeply for people with quite high anxiety. For an item like "I sometimes feel nervous," the curve would start rising much earlier, reflecting the fact that even moderately anxious people frequently endorse it.

Two features of the curve do most of the work. The location of the steep transition in the middle tells you where on the trait the item is most useful, which is $\beta$. The steepness of that transition tells you how sharply the item discriminates between people just below and just above that point, which is $\alpha$. The plot below shows a single item response curve to make these features concrete before we move into the models.

Below is a fake iterm resonse curve that serves as a demonstration.

```{r irc-intro, echo = F}
# Use LSAT item 3 as the basis for this illustration
# Item 3 has moderate difficulty and discrimination, representative of the dataset

p_item  <- mean(LSAT[, 3])           # empirical endorsement rate (p-value)
beta    <- qlogis(p_item)            # rough beta estimate
alpha   <- 1.5                       # arbitrary alpha for illustration

# setting up fake data
theta   <- seq(-4, 4, length.out = 200)
p       <- exp(alpha * (theta - beta)) / (1 + exp(alpha * (theta - beta)))
irc_df  <- data.frame(theta = theta, prob = p)

ggplot(irc_df, aes(x = theta, y = prob)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_vline(xintercept = beta, linetype = "dashed", color = "firebrick") +
  geom_hline(yintercept = 0.5,  linetype = "dashed", color = "firebrick") +
  annotate("text", x = beta + 0.15, y = 0.1,
           label = paste0("β = ", round(beta, 2)), color = "firebrick", hjust = 0) +
  labs(
    x        = expression(theta),
    y        = "P(X = 1 | θ)",
    title    = "A Single Item Response Curve",
    subtitle = "Dashed lines mark β, where P(endorsement) = .50"
  ) +
  theme_minimal()
```

The dashed lines mark $\beta = 0.21$, which is the point where the curve crosses the 50% probability line. Think of it as the item's tipping point: people to the left of that vertical line are more likely than not to get the item wrong, and people to the right are more likely than not to get it right. Since $\beta = 0.21$ is close to $\theta = 0$, this is a roughly average difficulty item. A person of average ability (recall that, since $\theta$ is standardized, $\theta = 0$ means the person has average competency) has about 50:50 chance of getting it correct. Notice also how the curve rises steeply right around the $\beta$ tipping point and flattens out toward the extremes. That steepness is $\alpha$ saying that the item is most useful for telling people apart right around the middle of the trait range, and much less useful for people who are very high or very low on the trait.

Everything that follows in this section is really just a description of how different IRT models handle the shape and positioning of this curve across a set of items.

### 2.4 The One-Parameter Logistic Model (1PL / Rasch)

The simplest IRT model constrains all discrimination parameters to the same value, usually 1. So items differ only in their location. This is often called a **1PL model** (short for *one-parameter logistic model*). Conceptually, this is like saying that every question on an exam is equally good at discriminating between students, and the only thing that varies is difficulty. Some questions are hard and some are easy, but all of them are assumed to do the same quality of job separating students who know the material from those who do not. In practice this is a strong assumption, and one that real data often violate.

Mathematically, the curve is described as

$$
P(X = 1 \mid \theta) = \frac{\exp(\theta - \beta)}{1 + \exp(\theta - \beta)}
$$

The only thing that varies across items is $\beta$, the location. As a result, all item response curves under the 1PL have exactly the same shape and steepness, just shifted left or right along the $\theta$ axis depending on how easy or hard the item is. An easy item has a curve shifted to the left, meaning even people with low trait levels have a reasonable chance of endorsing it. A hard item has a curve shifted to the right, meaning you need to be well above average on the trait before endorsement becomes likely.

Because all slopes are equal, item response curves for different items are parallel. They have the same shape and steepness, just shifted left or right along the $\theta$ axis. This can be understood better as item information, which is discussed in Section 3.

The 1PL (and its close relative the **Rasch model**) has appealing mathematical properties. Person and item parameters are separable, which in plain terms means the difficulty of an item does not depend on who you happened to test, and the estimate of a person's ability does not depend on which particular items they happened to receive. The raw total score is also a sufficient statistic for $\theta$, meaning that under the 1PL your score is entirely determined by how many items you got right, not which items you got right Get items 1, 3, and 5 correct or items 2, 4, and 5 correct and you will get the same $\theta$ estimate either way.

In practice, the equal-discrimination assumption is quite strong and often fails empirically. On an anxiety scale, an item like "I cannot leave my house because of fear" might discriminate very sharply at the high end of the trait, while an item like "I sometimes feel a little stressed" might be much more gradual and diffuse. Treating those two items as equally discriminating is a meaningful constraint that real data frequently reject. The 1PL is a useful baseline, but you should expect the 2PL to fit most real psychological data better, discussed next.

Below is a fake example IRC for three difficulty parameters for an "easy" item, a "moderate" item, and a "hard" item.

```{r irc-1pl, echo = F}
theta <- seq(-4, 4, length.out = 200)
beta_vals <- c(-1, 0, 1.5)
labels <- c("Easy (β = -1)", "Moderate (β = 0)", "Hard (β = 1.5)")

irc_1pl <- do.call(rbind, lapply(seq_along(beta_vals), function(i) {
  p <- exp(theta - beta_vals[i]) / (1 + exp(theta - beta_vals[i]))
  data.frame(theta = theta, prob = p, item = labels[i])
}))

ggplot(irc_1pl, aes(x = theta, y = prob, color = item)) +
  geom_line(linewidth = 1) +
  labs(x = expression(theta), y = "P(X = 1 | θ)",
       title = "1PL Item Response Curves", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Notice that the three curves are parallel with same shape and steepness. They are just shifted horizontally. That is the defining feature of the 1PL. Items are distinguished only by where they sit on the trait axis, not by how sharply they discriminate.

### 2.5 The Two-Parameter Logistic Model (2PL)

The 2PL frees the discrimination parameter to vary across items. Items now have both a slope and a location. Mathematically, the curve is described as

$$
P(X = 1 \mid \theta) = \frac{\exp(\alpha(\theta - \beta))}{1 + \exp(\alpha(\theta - \beta))}
$$ Unlike the 1PL, now the $\alpha$ term enters the function and allows each curve to differ in their slope and discrimination ability. A highly discriminating item is preferred over a less discriminating item for people near the item's location ($\beta$), but provides less information for people far from it.

An example of a fake 2PL plot is given below for a "high", a "moderate", and a "low" discriminatory item.

```{r irc-2pl, echo = F}
alpha_vals <- c(0.5, 1.0, 2.0)
beta_fixed <- 0
labels_a <- c("Low discrimination (α = 0.5)",
               "Moderate discrimination (α = 1.0)",
               "High discrimination (α = 2.0)")

irc_2pl <- do.call(rbind, lapply(seq_along(alpha_vals), function(i) {
  p <- exp(alpha_vals[i] * (theta - beta_fixed)) /
       (1 + exp(alpha_vals[i] * (theta - beta_fixed)))
  data.frame(theta = theta, prob = p, item = labels_a[i])
}))

ggplot(irc_2pl, aes(x = theta, y = prob, color = item)) +
  geom_line(linewidth = 1) +
  labs(x = expression(theta), y = "P(X = 1 | θ)",
       title = "2PL Item Response Curves: Effect of Discrimination", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

All three curves cross the .50 probability line at $\theta = 0$. The three items share the same location ($\beta = 0$), but notice how much steeper the high-discrimination curve is ($\alpha = 2$). A highly discriminating item draws a sharp line between people just below and just above the threshold. A low-discrimination item is gradual. It provides some information across a wide range of $\theta$, but very little at any particular point.

Here is how these two might in a questionnaire. An item with high $\alpha = 2$ and low $\beta = .2$ would discriminate well for individuals low in the trait. If you are building a test to measure patients' anxiety and you want to differentiate between people with no anxiety and some amount of anxiety, you would need an item with these values. Similarly, if you want an item that is good at discriminating high anxiety from debilitating anxiety, you would need an item like $\alpha = 2$ and $\beta = .9$.

It is much rarer when you would want items with low $\alpha$ discrimination. Some examples, however, might include situations where you are building a very short scale intended to give a rough sense of where someone falls across the entire trait range, rather than making precise distinctions at any particular point. A single broad item that provides modest information everywhere is sometimes more useful than a sharp item that is highly informative in one narrow band and nearly useless everywhere else. Low discrimination items also tend to appear in measures of fuzzy or multifaceted constructs like quality of life or general wellbeing, where the construct itself resists sharp boundaries and a gradual, diffuse item may actually be a more honest reflection of how the trait works in practice.

### 2.6 The 3PL and 4PL Models

In achievement testing, guessing is a real concern. Think about a multiple choice exam with four answer options. A student who has no idea what the correct answer is still has a 25% chance of getting it right just by picking randomly. This means that even for people at the very bottom of the ability distribution, the probability of a correct response never quite reaches zero. That nonzero floor is called the lower asymptote, and the standard 2PL has no way to represent it. If you look at the plot above, the 2PL curve just keeps dropping toward zero as ability decreases, which is unrealistic when blind guessing is possible.

The 3PL model handles this with a new "guessing" parameter called $c$. This is put into the probability curve in the following manner

$$
P(X = 1 \mid \theta) = c + (1 - c) \cdot \frac{\exp(\alpha(\theta - \beta))}{1 + \exp(\alpha(\theta - \beta))}
$$

The parameter $c$ describes the lower asymptote "guess" probability. For a four-option multiple-choice item, $c$ would plausibly be .25 since an individual would only have a $1/4$ chance of getting any answer correctly by just guessing.

The 4PL model adds an upper asymptote parameter $d$ for **inattention**. Think about a student who knows the material cold but misreads the question, runs out of time, or simply clicks the wrong answer by accident. For most items this is a minor concern, but for very easy items where nearly everyone should succeed, even a small rate of careless errors means the curve never quite reaches 1.0 at the top of the trait range. The 4PL adds an upper asymptote parameter to the probability curve function $d$ to capture this reality:

$$
P(X = 1 \mid \theta) = c + (1 - d) \cdot \frac{\exp(\alpha(\theta - \beta))}{1 + \exp(\alpha(\theta - \beta))}
$$

Both $c$ and $d$ are on the probability metric (bounded between 0 and 1). The 3PL is common in educational measurement but less relevant for personality and clinical research where guessing is minimal. For most psychological applications, the 2PL is the appropriate model.

Rarely would a psychologist introduce $d$, but one scenario where it becomes relevant is high-stakes clinical assessment, where a patient who is genuinely at the top of the anxiety distribution might still occasionally fail to endorse an item simply because the wording did not resonate with how they personally experience the symptom. Sometimes, it is important to keep potential false-negatives into account.

------------------------------------------------------------------------

## 3. Item and Test Information

### 3.1 Item Information Functions

One of the most useful features of IRT is that it tells you exactly how much information about $\theta$ each item provides and *where* along the trait continuum it provides that information. Think of information as measurement precision: a high information value at a particular $\theta$ means the item is doing a good job of pinning down exactly where on the trait a person sits. A low information value means the item is contributing little, essentially shrugging at that region of the scale. In practice this matters enormously. A clinical depression screener administered to a general population sample might have most of its items clustered in the moderate range of the trait, providing excellent precision for distinguishing mildly from moderately depressed people but doing almost nothing to differentiate someone with severe depression from someone who is completely debilitated. IRT makes that invisible inefficiency visible by showing you exactly where along the trait your test is sharp and where it is essentially guessing.

Information helps in this situation by turning a vague sense that your test might not work well at the extremes into a precise, actionable picture of exactly where it falls short and what you would need to fix it.

First I will go into some technical detail but try to bring this back to actionable advice. The item information function $I_i(\theta)$ quantifies this:

-   **1PL:** $I_i(\theta) = P_i(\theta) \cdot Q_i(\theta)$
-   **2PL:** $I_i(\theta) = \alpha_i^2 \cdot P_i(\theta) \cdot Q_i(\theta)$

In both cases $P_i(\theta)$ is the probability of endorsing the item at a given trait level and $Q_i(\theta) = 1 - P_i(\theta)$ is the probability of not endorsing it. Their product $P \cdot Q$ is highest when $P = .50$, meaning an item is most informative right at the point where endorsement becomes a coin flip. Move too far above or below the item's location and one of those probabilities approaches zero, dragging the product down with it. The 2PL adds $\alpha^2$ as a multiplier, which is why discrimination matters so much: an item with $\alpha = 2.0$ has $\alpha^2 = 4$, meaning it carries four times the information of an item with $\alpha = 1.0$ at the same location, even though both items have the same $P \cdot Q$ at their peak. Discrimination does not change where the item is most informative, it just determines how much information it delivers there.

Information is always highest near an item's location parameter and falls off on both sides. This makes sense as the item was built with that location in mind. A highly discriminating item produces a tall, narrow information spike, and a weakly discriminating item produces a low, broad hill.

An example of a fake information curve for three different items is given below for demonstration.

```{r item-info}
items <- list(
  list(alpha = 1.5, beta = -1,   label = "α=1.5, β=-1"),
  list(alpha = 2.0, beta =  0,   label = "α=2.0, β=0"),
  list(alpha = 0.8, beta =  1.5, label = "α=0.8, β=1.5")
)

info_data <- do.call(rbind, lapply(items, function(it) {
  p  <- exp(it$alpha * (theta - it$beta)) /
        (1 + exp(it$alpha * (theta - it$beta)))
  ii <- it$alpha^2 * p * (1 - p)
  data.frame(theta = theta, info = ii, item = it$label)
}))

ggplot(info_data, aes(x = theta, y = info, color = item)) +
  geom_line(linewidth = 1) +
  labs(x = expression(theta), y = "Item Information",
       title = "Item Information Functions", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The item with $\alpha = 2.0$ has the highest peak — nearly four times the height of a slope-1.0 item at the same location. The item with $\alpha = 0.8$ is spread out and low, indicating weak discrimination across the whole trait range.

The intuition behind the information formula connects directly back to trait discrimination, and understanding it has real practical value for anyone designing or evaluating a psychological measure.

Consider two people responding to the item "I have panic attacks every day." One has moderate anxiety, the other has severe anxiety. If this is a highly discriminating item, the moderately anxious person almost certainly says no and the severely anxious person almost certainly says yes. That clean separation is what makes the response informative: a clinician who knows how a patient answered that item has learned something meaningful about where they sit on the anxiety continuum. Now imagine a poorly worded version of the same item where both groups have about a 50/50 chance of endorsing it regardless of their actual severity. The response still comes in, but it tells you almost nothing. You administered an item, used up a patient's time and attention (not to mention your precious grant budget), and walked away knowing barely more than you did before.

This is where item information becomes a practical tool rather than just a statistical concept. Suppose you are a psychologist working in a hospital and you are evaluating whether your current depression screener is fit for purpose. The information function might reveal that your 20-item scale delivers excellent precision for patients in the mild to moderate range but essentially loses its ability to differentiate among patients at the severe end, which is exactly where your clinical decisions are most consequential. That tells you something actionable: you do not need more items, you need better-targeted items. Specifically, you need items with high discrimination and high location parameters, items that only start to differentiate once a patient is already quite severely depressed. Without the information function, you would never know this from a reliability coefficient alone. Alpha might be .90 and you would feel confident in your scale, all while missing the patients who need the most precise assessment.

### 3.2 Test Information and the Standard Error of Measurement

Once you have item information functions for each item in your scale, combining them can give you the overall information of your test. I.e., how well your test informs you about the latent psychological trait. To do this is straightforward: you simply add them up. The resulting **test information function** tells you the total measurement precision your scale delivers at every point along the trait continuum. Think of it like stacking the contributions of each item on top of one another. An item that is highly informative near $\theta = 0$ adds a tall peak to that region of the test information curve. An item targeted at $\theta = 2$ adds its contribution further up the scale. The final curve reflects the collective measurement capacity of the whole instrument, and its shape tells you immediately where your test is doing its best work and where it is running thin.

The **standard error of measurement** follows directly from test information and is its practical translation. High test information at a given $\theta$ means low standard error, meaning your score estimates of the latent trait are precise in that region. Low test information means high standard error, meaning your estimates of the latent trait are uncertain. This is the IRT answer to the CTT limitation we discussed in Section 1: rather than a single SEM that gets applied uniformly to everyone, you now have a standard error that varies across the trait and honestly reflects the fact that your test measures some people more precisely than others.

Mathematically, test information is the sum of item information functions:

$$TINFO(\theta) = \sum_{i=1}^{n} I_i(\theta)$$

And the standard error of measurement at any given trait level is:

$$SE(\hat{\theta}) = \frac{1}{\sqrt{TINFO(\theta)}}$$ The standard error of measurement varies across $\theta$. A test with items clustered around $\beta = 0$ is highly precise for people near the population mean but much less precise for people with extreme trait values. A clinical screening tool might deliberately place items at higher $\beta$ values to be most precise in the range where diagnostic decisions are made. It all depends on the purpose of your test and what population you are targeting.

An example plot for test information and standard error of measurement is given below for a fake test

```{r test-info, echo =F}
tinfo <- aggregate(info ~ theta, data = info_data, FUN = sum)
tinfo$se <- 1 / sqrt(tinfo$info)
p_info <- ggplot(tinfo, aes(x = theta, y = info)) +
  geom_line(linewidth = 1, color = "steelblue") +
  labs(x = expression(theta), y = "Test Information",
       title = "Test Information Function") +
  theme_minimal()

p_se <- ggplot(tinfo, aes(x = theta, y = se)) +
  geom_line(linewidth = 1, color = "firebrick") +
  labs(x = expression(theta), y = "Standard Error",
       title = "Standard Error of Measurement") +
  theme_minimal()

print(p_info)
print(p_se)
```

As one can see, the standard error of measurement curve is the mirror image of the information curve. Where we can see information in the test is high, the standard error is low. And where the test information is low, there is much more error. This can inform us that the test measures most precisely where it has the most items with high discrimination located.

In clinical psychology, this has direct implications for how you design and evaluate assessment tools. A depression screener intended for use in a primary care setting, where the goal is to catch patients who might be slipping into clinical range, should have its information peak somewhere in the mild to moderate region of the trait. If the test information curve shows a peak at $\theta = 0$ but drops off sharply above $\theta = 1.5$, a clinician using it to identify severely depressed patients is essentially working with a blunt instrument when they need a sharp knife. The standard error curve will show this honestly. Large standard errors at the high end of the trait mean that two patients who are genuinely quite different in their depression severity (one who struggles to get out of bed most days and one who is completely unable to function) might receive scores so close together that the test cannot distinguish between them. For a clinician trying to make decisions about level of care or treatment intensity, that is a meaningful failure.

### 3.3 Marginal Reliability

The concepts of reliability in item response theory (IRT) and classical test theory (CTT) are more connected than they might initially appear. Recall from Section 1 that CTT reliability is a single number summarizing how much of the observed score variance is attributable to true score variance. That number is useful but it collapses everything into one figure and applies it uniformly to everyone. Marginal reliability is the IRT equivalent: it asks what the average reliability of the test is across the full distribution of $\theta$ in the population, integrating the information function over where people actually tend to fall on the trait. This is mathematically given as

$$\bar{\rho} = 1 - \frac{1}{\overline{TINFO}}$$

Where $\overline{TINFO}$ is the average test information across the population distribution of $\theta$. The formula has an intuitive structure: when average test information is high, the fraction $\frac{1}{\overline{TINFO}}$ is small, and marginal reliability approaches 1. When test information is low on average, that fraction grows and reliability drops. In this sense marginal reliability is doing exactly what CTT reliability does, summarizing overall precision as a single number between 0 and 1, but it is built from the information function rather than from observed score variances.

To make this concrete: imagine a research team develops a social anxiety scale by piloting it on a clinical sample of patients seeking treatment for anxiety disorders. Most of the participants in that pilot sample have moderate to severe social anxiety, so the items end up being calibrated toward the higher end of the trait. The scale has good marginal reliability in that clinical sample, perhaps around .90, and the team publishes it feeling confident.

A few years later, a different research group wants to use the same scale in a large community survey where most participants have little to no social anxiety. When they compute marginal reliability in their sample, it is noticeably lower, perhaps around .72, even though they are using the exact same items. The CTT reliability coefficient would tell them something went wrong but gives no indication of why. The test information function tells them immediately: the items are delivering their precision up at $\theta = 1.5$ and above, but the bulk of their community sample is sitting near $\theta = -0.5$, a region where the scale is barely measuring anything. The scale was not poorly designed, it was just designed for a different population than the one they are using it with.

The researchers would get to this conclusion by plotting the test information curve alongside the distribution of $\theta$ estimates from their sample. The information curve shows where the test is precise. The $\theta$ distribution shows where their participants actually are. When you overlay the two, the mismatch becomes immediately obvious: a tall information peak sitting at $\theta = 1.5$ with most of the sample clustered near $\theta = -0.5$ tells the whole story at a glance. The region where the test is working hardest is almost entirely unpopulated, and the region where most participants sit is where the information curve has already flattened out near zero.

The bottom line is this: you cannot simply take a test developed for one context and apply it to another. The population you are measuring matters just as much as the items themselves.

------------------------------------------------------------------------

## 4. Assumptions and Model Checking

Before fitting any IRT model, it is worth verifying that the data meet the core assumptions. There are two, and they are two sides of the same coin.

**Unidimensionality:** The items should all reflect a single underlying latent trait. If items have substantial secondary dimensions, a single $\theta$ cannot fully account for item covariances, and the model will be misspecified.

**Local independence:** After conditioning on $\theta$, responses to different items should be uncorrelated. This follows directly from unidimensionality: if $\theta$ is the only driver of responses, knowing how someone answered item 3 adds no information about item 7 once $\theta$ is known. Violations typically indicate that items share specific content beyond the common dimension, or that the scale is genuinely multidimensional.

### 4.1 Unidimensionality

The most common approach is parallel analysis or inspection of the scree plot from a principal components analysis of the tetrachoric correlation matrix. Strong unidimensionality is indicated by a dominant first factor with a large eigenvalue and a sharp drop-off before the second factor. A ratio of the first to second eigenvalue of 3:1 or greater is generally taken as evidence of sufficient unidimensionality for IRT.

```{r parallel-analysis}
# Compute tetrachoric correlations and inspect eigenvalues
tet_cor <- psych::polychoric(LSAT)$rho
eigs    <- eigen(tet_cor)$values

cat("Eigenvalues:\n")
print(round(eigs, 3))
cat("\nRatio (EV1 / EV2):", round(eigs[1] / eigs[2], 2), "\n")
```

For the LSAT, you should see a dominant first eigenvalue with a ratio well above 3:1, consistent with a unidimensional structure for a test explicitly designed to measure a single ability construct.

### 4.2 Local Independence and the Q3 Statistic

The Q3 statistic (Chen & Thissen, 1997) is the correlation between the residuals of two items after removing the effect of $\theta$. If local independence holds, these residual correlations should be near zero. Values above about .20 flag item pairs that are more correlated than the model accounts for, suggesting a local dependency.

```{r q3}
fit_2pl_lsat <- mirt(LSAT, 1, itemtype = "2PL", verbose = FALSE)
q3_mat <- residuals(fit_2pl_lsat, type = "Q3")
print(round(q3_mat, 3))
```

Values near zero across all pairs indicate the 2PL is capturing the shared variance among items. Elevated Q3 values for particular pairs warrant investigation: do those two items share specific content beyond the broader construct?

### 4.3 Monotonicity via Mokken Scaling

Before committing to a parametric IRT model, it is worth checking whether item response functions are monotonically increasing. Monotonicity — the assumption that higher $\theta$ leads to higher probability of endorsement — is a prerequisite for any sensible IRT analysis. The `mokken` package provides a nonparametric check.

```{r mokken}
mono_check <- check.monotonicity(LSAT)
summary(mono_check)
```

The summary flags any items with significant violations. For a well-constructed scale, violations are uncommon but do occur — particularly for items with poor face validity or those interpreted differently at different trait levels. Items with clear violations should be reviewed before fitting parametric models.

------------------------------------------------------------------------

## 5. Fitting Dichotomous IRT Models in R

We now have the conceptual foundation to fit models to real data. The `mirt` package is the standard tool for IRT in R — it handles dichotomous and polytomous models, produces IRT-parameterized output, computes fit indices, and generates all the plots we have discussed.

### 5.1 Fitting the Models

```{r fit-models}
fit_1pl <- mirt(LSAT, 1, itemtype = "Rasch", verbose = FALSE)
fit_2pl <- mirt(LSAT, 1, itemtype = "2PL",   verbose = FALSE)
```

The first argument is the data matrix, the second is the number of factors (1 for unidimensional IRT), and `itemtype` specifies the model. `"Rasch"` constrains all slopes to 1; `"2PL"` estimates them freely.

### 5.2 Extracting Parameters

By default, `mirt` returns parameters in intercept-slope form (`a` = slope, `d` = intercept where $d = \gamma = -\alpha\beta$). Setting `IRTpars = TRUE` converts these to the more interpretable slope-location form:

```{r params}
# IRT-parameterized: a = discrimination, b = location
coef(fit_2pl, IRTpars = TRUE, simplify = TRUE)$items
```

The `a` column is the discrimination parameter $\alpha$; the `b` column is the location $\beta$. For the LSAT you should see discriminations ranging from moderate to relatively high, and locations spread across the $\theta$ range — reflecting the variation in item difficulty that makes the LSAT a useful measuring instrument.

The intercept form is available by dropping `IRTpars = TRUE`, which is sometimes needed for downstream calculations:

```{r params-intercept}
coef(fit_2pl, IRTpars = FALSE, simplify = TRUE)$items
```

### 5.3 Comparing Model Fit

A central question is whether freeing the discrimination parameters — moving from 1PL to 2PL — meaningfully improves fit. We address this with absolute fit indices and with a likelihood ratio test.

```{r fit-indices}
# Absolute fit: M2 statistic, RMSEA, CFI
# Good fit: non-significant M2, RMSEA < .05, CFI > .95
M2(fit_1pl)
M2(fit_2pl)

# Relative fit: likelihood ratio test
anova(fit_1pl, fit_2pl)
```

In research with the Taylor Manifest Anxiety Scale (25 items), moving from a 1PL to the 2PL produced a substantial improvement in fit, with item discriminations ranging from about 0.45 to 2.92 — clearly inconsistent with the equal-slopes assumption. The M2 statistic dropped considerably and RMSEA fell to acceptable levels. Most psychological scales show similar patterns: items really do discriminate at different rates, and the 2PL captures that variation.

Adding a guessing parameter (3PL) often improves $\chi^2$-based fit but not AIC or BIC, and it introduces estimation challenges. For psychological scales rather than achievement tests, the 2PL is usually the appropriate stopping point.

### 5.4 Visualizing Results

`mirt` produces a clean set of diagnostic plots with a single `plot()` call.

```{r item-plots}
# Item trace plots: one IRC per item
plot(fit_2pl, type = "trace")

# Item information functions
plot(fit_2pl, type = "infotrace")

# Test information and SE
plot(fit_2pl, type = "infoSE")

# Marginal reliability as a function of theta
plot(fit_2pl, type = "rxx")
```

When reviewing these plots, ask: are the item response curves well-behaved? Does the test information peak in the region of $\theta$ most relevant to the population you care about? Are there items with flat information functions near zero across the full trait range? Those items are candidates for revision or removal.

------------------------------------------------------------------------

## 6. The Graded Response Model (GRM)

### 6.1 From Dichotomous to Polytomous

Most psychological scales do not use dichotomous items. Likert-type response formats are far more common. These polytomous items require a different IRT model, because a 5-point item cannot be scored with a single logistic curve.

The conceptual bridge is straightforward: a $K$-category item is equivalent to $K - 1$ cumulative dichotomies. A 5-point item with categories 1 through 5 can be decomposed into four nested comparisons: (category 1) vs. (categories 2–5); (categories 1–2) vs. (categories 3–5); (categories 1–3) vs. (categories 4–5); and (categories 1–4) vs. (category 5). Each is a dichotomous decision — "below this threshold" vs. "at or above this threshold" — and each can be modeled with a logistic curve.

**The data: BFPT-E.** The GRM examples use the Extraversion subscale of the Big Five Personality Test (BFPT-E): 10 items on a 5-point format, $N = 2{,}000$. Items reflect different facets of extraversion — sociability, assertiveness, positive affect — and cover a range of threshold locations and discrimination values. Estimated slopes under the GRM range from about 1.40 to 2.58, and threshold parameters span beyond $\pm 2$ for some items, providing good coverage of the trait continuum.

### 6.2 Threshold Response Curves

The Graded Response Model (Samejima, 1969) models each of the $K - 1$ cumulative dichotomies with a 2PL-like curve called a Threshold Response Curve (TRC). The probability of responding *above* boundary $v$ for item $i$ is:

$$
P^*_{ix}(\theta) = \frac{\exp[\alpha_i(\theta - \beta_{iv})]}{1 + \exp[\alpha_i(\theta - \beta_{iv})]}
$$

where $\alpha_i$ is the item's slope (shared across all boundaries) and $\beta_{iv}$ is the location of boundary $v$. The **homogeneous constraint** is the defining feature of the GRM: all TRCs for a given item share the same slope. Only the threshold locations differ.

The threshold parameters must be ordered: $\beta_{i1} < \beta_{i2} < \cdots < \beta_{i,K-1}$. This ordering reflects the cumulative nature of the response process. The threshold $\beta_{iv}$ has a concrete interpretation: it is the value of $\theta$ at which the probability of responding in category $v$ or higher equals .50.

### 6.3 Category Response Curves

The TRCs give cumulative probabilities. To get the probability of being in a *specific* category, you take the difference between adjacent TRCs:

$$
P(X = k \mid \theta) = P^*_{i,k-1}(\theta) - P^*_{i,k}(\theta)
$$

with boundary conditions $P^*(0) = 1$ and $P^*(K) = 0$. These are the Category Response Curves (CRCs).

The shape of CRCs is diagnostic. The lowest category has a monotonically decreasing CRC — only people with very low $\theta$ choose it. The highest category has a monotonically increasing CRC. All middle categories produce unimodal, bell-shaped CRCs that peak somewhere in the middle of the trait range. Widely spaced thresholds produce well-separated, narrow CRCs. Closely spaced thresholds produce overlapping, flat CRCs that never dominate the response space — a sign that the response scale is too fine for how people are actually using it.

Finally, summing the expected category scores across $\theta$ gives the Item Response Curve (IRC), which plots the expected item score as a function of trait level. Summing IRCs across items yields the Scale Response Curve (SRC), the polytomous analog of the test characteristic curve.

### 6.4 Fitting the GRM in R

```{r grm-fit}
# Load the BFPT-E data (replace path when data is available)
# bfpt_e <- read.csv("data/bfpt_e.csv")

# Fit the GRM
# GRM_fit <- mirt(bfpt_e, 1, itemtype = "graded", verbose = FALSE)

# IRT-parameterized: a = slope, b1-b4 = threshold locations
# coef(GRM_fit, IRTpars = TRUE, simplify = TRUE)$items

# Intercept form: a = slope, d1-d4 = intercepts
# coef(GRM_fit, IRTpars = FALSE, simplify = TRUE)$items

# Column means across items
# apply(coef(GRM_fit, IRTpars = TRUE, simplify = TRUE)$items, 2, mean)
```

### 6.5 Interpreting GRM Parameters

Each item in a GRM solution has one slope and $K - 1$ threshold parameters. For the BFPT-E:

Slopes ranged from 1.40 to 2.58. Higher slopes indicate sharper, better-separated CRCs — items with slopes above 2.0 are highly discriminating. Thresholds ($\beta_1$ through $\beta_4$) represent the four cumulative boundaries on the $\theta$ scale. For a well-functioning item, thresholds should be spread across a reasonable range and in the expected order. An item with all four thresholds clustered near $\theta = -0.5$ is essentially only informative for people near the low end of the trait.

There is an important relationship between slope and threshold spacing. In intercept parameterization, $\beta_{iv} = -\gamma_{iv} / \alpha_i$. This means that when you re-express intercepts as threshold locations, higher slopes *compress* the apparent threshold spacing on the $\theta$ scale. Two items can have the same intercepts but different slopes, and the one with the higher slope will show more tightly clustered thresholds. The thresholds are not independent of the slope.

### 6.6 Item Information in Polytomous Models

Polytomous items generally carry more total information than dichotomous items, and they spread that information more broadly across the trait range. A 5-point item is doing more measurement work than a 0/1 item — it is making four distinct discriminations rather than one.

```{r grm-info}
# plot(GRM_fit, type = "infotrace")  # item information per item
# plot(GRM_fit, type = "infoSE")     # test information and SE
# plot(GRM_fit, type = "trace")      # CRCs for each item
```

Item information plots from a GRM often have multiple local peaks — one near each threshold location. The item is most discriminating near each cumulative boundary, not just at a single point.

------------------------------------------------------------------------

## 7. The Nominal Response Model (NRM)

### 7.1 The NRM Equation

The Graded Response Model assumes that the response categories have a known, fixed ordering. This is reasonable for Likert-format items where categories are explicitly ordered by intensity. But what if you are not confident the categories are being used in their intended order? Or what if the categories are genuinely unordered, as in multiple-choice items where wrong answers might carry information?

The Nominal Response Model (Bock, 1972) handles these cases. It is the most general and least constrained of the common polytomous IRT models — every response category gets its own slope ($\alpha_{ix}$) and its own intercept ($c_{ix}$):

$$
P_{ix}(\theta) = \frac{\exp(\alpha_{ix}\theta + c_{ix})}{\sum_{x=0}^{K_i - 1} \exp(\alpha_{ix}\theta + c_{ix})}
$$

This is a divide-by-total model: the probability of each category is its exponential weight divided by the sum of all exponential weights. For identification, either the slopes and intercepts are constrained to sum to zero across categories, or the parameters for the lowest category are fixed to zero.

The category with the most positive $\alpha_{ix}$ will have a monotonically increasing CRC — people with higher $\theta$ increasingly prefer that category. The category with the most negative $\alpha_{ix}$ will have a monotonically decreasing CRC. If the slope ordering matches the intended category ordering, the data are consistent with ordered categories — this is the empirical check the NRM provides.

**The data: V-HEM.** The NRM examples use the Veterans Healthcare Engagement Measure, a 23-item scale measuring engagement with healthcare in veteran populations. Items are rated on a 5-category format ("Not at all true" to "Very true"). Category 0 was endorsed very rarely, so categories 0 and 1 were collapsed, yielding 4-category scoring (0 through 3) for analysis. This kind of category collapsing is common when response scale extremes are almost never used.

### 7.2 Transformed Parameters: CBD and Category Intersection

The raw NRM slopes are on an arbitrary scale that is difficult to interpret directly. Two derived quantities are more useful.

The **Category Boundary Discrimination (CBD)** for boundary $j$ is the difference between adjacent slopes:

$$\text{CBD}_j = \alpha_{i,j} - \alpha_{i,j-1}$$

This is analogous to the GRM threshold response curve slope — it tells you how sharply the model discriminates between adjacent categories at their intersection. If all CBDs are positive and increasing, the categories are functioning in their intended order. For the V-HEM, CBDs were all positive (means approximately 0.64, 1.02, and 1.70 across categories), confirming that the four response categories were functioning as intended.

The **Category Intersection** $\delta_j$ is the value of $\theta$ at which adjacent CRCs cross:

$$\delta_j = \frac{c_{i,j-1} - c_{i,j}}{\alpha_{i,j} - \alpha_{i,j-1}}$$

This is the threshold on the $\theta$ scale where the model transitions from predicting category $j - 1$ to predicting category $j$. Well-ordered intersections spanning the trait range indicate that all categories are being used meaningfully.

### 7.3 Fitting the NRM in R

```{r nrm-fit}
# vhem <- read.csv("data/vhem.csv")

# Collapse categories 0 and 1 due to low endorsement of category 0
# vhem_collapsed <- as.data.frame(
#   lapply(vhem, function(x) ifelse(x == 0, 1, x) - 1)
# )

# Fit the NRM
# NRM_fit <- mirt(vhem_collapsed, 1, itemtype = "nominal", verbose = FALSE)
# coef(NRM_fit, simplify = TRUE)$items
```

### 7.4 Interpreting NRM Output

From the V-HEM analysis, item slopes ($a^*$) ranged from approximately 0.48 to 1.79, with a mean around 1.12. Item 5 ("easy to refill medications") had a slope below 0.5 and very compressed CRCs — a weak item that barely discriminates anywhere along the trait range. Item 21 showed well-separated CRCs and high information, particularly in the moderate-to-high $\theta$ region.

Scoring coefficients reflected the IRT-optimal weighting for each response category. On average, coefficients increased monotonically across categories (means approximately 0, 0.57, 1.48, 3.0), consistent with the intended ordering. However, individual items showed irregularities — categories that are nominally ordered but whose scoring coefficients are not strictly increasing. These items may benefit from response category recoding or content revision.

EAP estimates from the fitted model were approximately normally distributed, and the relationship between total raw scores and EAP estimates was monotonically increasing but nonlinear. People with low raw scores are estimated to have lower $\theta$ than integer scoring would imply; people with high raw scores are estimated higher. This is the consequence of the NRM using IRT-optimal scoring weights rather than treating all category increments as equal.

------------------------------------------------------------------------

## 8. Computerized Adaptive Testing (CAT)

### 8.1 Why CAT?

Fixed-length tests are inefficient by design. A 40-item depression scale administered to everyone will include many items that are essentially uninformative for any given respondent. For someone with very high depression, easy items — those with low location parameters — offer almost no information: the probability of endorsement is near 1.0 regardless of exactly where on the $\theta$ scale they sit. For someone with low depression, the hard items are equally useless. Both types of items get administered anyway because the test is fixed-length and the same for everyone.

Computerized adaptive testing (CAT) eliminates this waste. Rather than administering the same items to all examinees, a CAT selects, in real time, the item from the available bank that provides the most information about the examinee's current estimated trait level. As more responses accumulate and the estimate sharpens, the algorithm homes in on a narrower region of the trait continuum, selecting items targeted to that region.

CAT rests entirely on IRT's person-invariance property: a person's $\theta$ does not depend on which particular items are used to estimate it. If the trait estimate were test-dependent, as it is in CTT, there would be no principled way to compare scores from adaptive tests that administered different items.

### 8.2 The CAT Algorithm

The basic CAT algorithm loops through five steps until a stopping criterion is met:

1.  **Initialize.** Assume $\hat{\theta} = 0$ and administer the item of median difficulty from the bank. Alternatives include starting at a specific prior $\hat{\theta}$ or randomly sampling from moderate-difficulty items to prevent overexposure of the same starting item.
2.  **Score the item.** Record the response.
3.  **Estimate** $\theta$. Update $\hat{\theta}$ using the accumulated response pattern.
4.  **Select the next item.** Choose the item from the remaining bank that provides the most information at the current $\hat{\theta}$, or that most reduces posterior SE under a Bayesian criterion.
5.  **Check the termination criterion.** If satisfied, stop and report $\hat{\theta}$. If not, administer the selected item and loop.

An important practical concern is **content constraints**. Pure maximum-information item selection tends to fixate on a small set of highly discriminating items, leading to overexposure and tests that feel narrow. Operational CATs impose constraints: no more than one item per content domain, no item administered more than some fraction of the time, etc.

### 8.3 Theta Estimation in CAT

Three estimation methods are used in practice:

**Maximum likelihood (ML):** Returns the $\theta$ that maximizes the likelihood of the observed response pattern. Unbiased, but undefined until at least one correct and one incorrect response has been observed — the likelihood function is monotone otherwise. ML requires EAP or MAP as a fallback at the start.

**Expected a posteriori (EAP):** Uses a prior distribution on $\theta$ (typically $N(0, 1)$) and returns the mean of the posterior. Biased — it shrinks estimates toward the prior mean — but more efficient than ML (smaller SE for the same number of items) and always defined. EAP is the most common choice in operational CAT.

**Maximum a posteriori (MAP):** Also Bayesian, returns the posterior mode rather than the mean. Less commonly used than EAP.

### 8.4 Item Selection and Termination

Under ML estimation, item selection at each step is simple: choose the item from the unused bank with the highest Fisher information at the current $\hat{\theta}$. Under Bayesian estimation, choose the item that most reduces posterior standard error.

For termination, two approaches are common. **Fixed-length** administers a predetermined number of items and stops — simple and administratively predictable, but it does not equalize precision across examinees. The **SE threshold** approach continues until \$SE(\hat{\theta}) \< \$ some criterion, equalizing precision but producing variable test length. In practice the SE criterion often cannot be met even after the full bank is exhausted, particularly for examinees at the extremes of the trait distribution.

### 8.5 A Worked Example: Likelihood Updating

To make the algorithm concrete, consider a 9-item bank where all slopes are $\alpha = 1.7$ and locations are evenly spaced from $-2.0$ to $+2.0$ in steps of 0.5. An examinee begins at item 5 ($\beta = 0.0$) and produces the response pattern 1, 1, 0, 1 across items 5, 6, 7, and 8.

```{r cat-likelihood}
theta_grid <- seq(-4, 4, length.out = 500)
alpha_val  <- 1.7
betas      <- c(0.0, 0.5, 1.0, 1.5)
responses  <- c(1,   1,   0,   1)

# Log-likelihood contribution of each item
log_lik_mat <- sapply(seq_along(betas), function(j) {
  p <- exp(alpha_val * (theta_grid - betas[j])) /
       (1 + exp(alpha_val * (theta_grid - betas[j])))
  if (responses[j] == 1) log(p) else log(1 - p)
})

# Cumulative normalized likelihood after each item
cum_lik <- do.call(rbind, lapply(1:4, function(j) {
  cl <- exp(rowSums(log_lik_mat[, 1:j, drop = FALSE]))
  cl <- cl / max(cl)
  data.frame(
    theta = theta_grid,
    lik   = cl,
    stage = paste0("After Item ", j + 4,
                   " (β=", betas[j], ", X=", responses[j], ")")
  )
}))

ggplot(cum_lik, aes(x = theta, y = lik)) +
  geom_line(color = "steelblue", linewidth = 1) +
  facet_wrap(~ stage, ncol = 2) +
  labs(x = expression(theta), y = "Normalized Likelihood",
       title = "Likelihood Updating Across CAT Items") +
  theme_minimal()
```

After item 5 (correct), the likelihood is a broad curve shifted right of center — weak evidence that the examinee is above average. After item 6 (also correct), the likelihood sharpens and shifts further right. After item 7 (incorrect), the likelihood is pulled back left and narrows, now indicating the examinee is above average but not by much. After item 8 (correct), the likelihood narrows further and the ML estimate settles near $\hat{\theta} \approx 1.2$. With each item, the curve both *moves* (updating the estimate) and *narrows* (reducing uncertainty).

### 8.6 CAT Efficiency: Applied Examples

The efficiency gains from CAT are well documented. In aptitude testing, CAT typically achieves the same precision as a full fixed-length test using about 50% of the items. Results in personality and clinical research vary by instrument design.

The **Absorption scale** (34 dichotomous items) produced a fixed-length CAT of 17 items — exactly 50% savings — with a correlation of .97 between full-scale $\theta$ and CAT $\hat{\theta}$.

The **Clinical Decision CAT** illustrates a different design goal: determine whether a patient's $\hat{\theta}$ falls above or below a clinical cut-point ($\theta = 2.0$), stopping when the 90% CI for $\hat{\theta}$ excludes that cut-point. In a simulation of $N = 1{,}000$ examinees, 85.1% required only 7 items out of 34, and classification was perfect — zero misclassifications.

The **NEO PI-R Anxiety (N1) facet** (8 items, 5-point GRM) reached a correlation of .93–.95 with full-scale $\theta$ after just 3–4 items. Item 3 (slope $= 2.06$, the highest-discriminating item) was administered first to virtually all examinees, as it provided the most information near $\theta = 0$.

### 8.7 Limits of CAT in Personality and Health Research

CAT is not a universal solution. Several features of personality and clinical measurement limit its applicability relative to achievement testing.

**Finite item pools.** Aptitude constructs have essentially unlimited potential items — any new math problem or reading passage is a candidate. Personality and psychopathology constructs do not. There are only so many non-redundant ways to ask about social anxiety or depression, and item banks tend to be small.

**Polytomous items reduce the efficiency advantage.** A well-designed polytomous item already spreads information across multiple regions of the trait continuum via its multiple thresholds. There may be less inefficiency for the CAT algorithm to eliminate — in well-designed polytomous scales, there is simply nothing for the algorithm to adapt to.

**Construct narrowness.** High item discriminations in personality research often reflect highly similar item content. A bank of nearly identically worded items will produce high $\alpha$ values, but those items are largely redundant. CAT selects for high information, which can mean administering conceptually repetitive items — statistically satisfying but potentially transparent or frustrating for respondents.

------------------------------------------------------------------------

## 9. Differential Item Functioning (DIF)

### 9.1 Measurement Invariance and Its Violations

For any psychological measure to support valid group comparisons, test scores must mean the same thing across groups. If a depression scale produces higher scores in women than in men, we want to be confident that the difference reflects a real difference in depression, not a difference in how the items function. This is the principle of measurement invariance.

In IRT terms, measurement invariance holds when an item has the same item characteristic curve for all groups. Given the same $\theta$, members of all groups should have the same probability of endorsement. When this condition fails — when an item is systematically easier or harder for one group at the same trait level — the item exhibits **differential item functioning (DIF)**. DIF can bias total scores and distort group mean comparisons in ways that are invisible to CTT reliability analyses.

Measurement invariance can also be assessed at the scale level by comparing test response curves across groups. A test with no DIF should produce identical test response curves in all groups.

### 9.2 Uniform vs. Non-Uniform DIF

DIF comes in two varieties, defined by the pattern of group differences across the trait continuum.

**Uniform DIF** occurs when item slopes are the same across groups but location parameters differ. One group needs to be at a systematically higher or lower trait level to achieve the same probability of endorsement. On the ICC plot, the two group curves are parallel — same shape, same steepness — but horizontally offset. The group disadvantage is constant across all values of $\theta$.

**Non-uniform DIF** occurs when item slopes differ across groups. The ICC curves have different steepnesses and will cross somewhere along the $\theta$ axis. The direction of the group difference reverses: one group may be advantaged at low $\theta$ and disadvantaged at high $\theta$, or vice versa. Non-uniform DIF is more complex to interpret and requires that slope differences are substantively meaningful, not just sampling noise.

### 9.3 Detecting DIF: Logistic Regression

The logistic regression approach (Swaminathan & Rogers, 1990) uses the total raw score $X$ as a proxy for $\theta$ and detects both types of DIF through a single model:

$$Z = b_0 + b_1(X) + b_2(G) + b_3(XG)$$

where $G$ codes group membership (0 = reference, 1 = focal) and $XG$ is the interaction. The parameters are interpreted as follows. $b_1$ captures the association between total score and item endorsement (ability matching). $b_2$ is the group mean difference on the item after controlling for ability — the uniform DIF coefficient. $b_3$ is the difference in slopes between groups — the non-uniform DIF coefficient.

```{r dif-logreg}
# Logistic regression DIF detection using the LSAT.
# We use a simulated grouping variable for illustration only.
set.seed(42)
group       <- sample(0:1, nrow(LSAT), replace = TRUE)
total_score <- rowSums(LSAT)
item_resp   <- LSAT[, 1]

dif_full    <- glm(item_resp ~ total_score + group + total_score:group,
                   family = binomial)
dif_uniform <- glm(item_resp ~ total_score + group,
                   family = binomial)
dif_null    <- glm(item_resp ~ total_score,
                   family = binomial)

# Test for uniform DIF (b2): does adding group improve fit?
anova(dif_null, dif_uniform, test = "LRT")

# Test for non-uniform DIF (b3): does the interaction improve fit?
anova(dif_uniform, dif_full, test = "LRT")

summary(dif_full)
```

Significance of the group term ($b_2$) indicates uniform DIF; significance of the interaction ($b_3$) indicates non-uniform DIF. In practice, you run this for every item in the scale and apply a correction for multiple testing (Bonferroni or false discovery rate).

### 9.4 Detecting DIF: IRT Likelihood Ratio Approach

The IRT-based approach is more principled because it uses $\theta$ rather than the raw score as the ability control. The key challenge is that you cannot simply calibrate parameters separately in each group and compare them — if groups have different mean $\theta$ values, the parameters will be on different scales and direct comparison is invalid.

The solution is **iterative linking** with anchor items. The procedure identifies items believed to be free of DIF and uses them to establish a common scale across groups. The focal item is then freed to have different parameters in each group, and a likelihood ratio test determines whether the freed parameters improve fit.

In the **model comparison approach**: the constrained model estimates all items with the same parameters in both groups (equivalent to single-group estimation). The unconstrained model frees the slope and location of one focal item while all other items serve as anchors. The $-2$ log-likelihood difference follows a $\chi^2$ distribution with degrees of freedom equal to the number of freed parameters (2 for the 2PL). A significant result indicates DIF.

```{r dif-irt}
# IRT-based DIF using mirt's multipleGroup() function.
group_factor <- factor(group)

# Constrained baseline: all parameters equal across groups
# fit_constrained <- multipleGroup(
#   LSAT, 1, group = group_factor,
#   invariance = c("slopes", "intercepts", "free_means", "free_var")
# )

# Free item 1: estimate its slope and intercept separately per group
# fit_free_item1 <- multipleGroup(
#   LSAT, 1, group = group_factor,
#   invariance = names(LSAT)[-1]   # anchor all items except item 1
# )

# Likelihood ratio test
# anova(fit_constrained, fit_free_item1)
```

A practical complication is that you need to identify DIF-free anchor items before you start. In practice, you often begin by assuming all items are anchors, identify those showing significant DIF, remove them from the anchor set, and repeat until convergence — this is iterative purification.

### 9.5 DIF and Multidimensionality

There is a deeper connection between DIF and multidimensionality worth understanding. IRT assumes unidimensionality but is routinely applied to scales that are merely "unidimensional enough" — scales with a strong common factor but also detectable secondary dimensions. The central irony is that differential group mean differences on those secondary dimensions are often the primary cause of DIF.

The MPQ Stress Reaction Scale illustrates this clearly. The scale has 23 items covering a broad stress reaction construct, but factor analysis reveals five correlated content subdimensions: tends to worry; nervous and tense; sensitive and vulnerable; easily upset and irritable; often feels guilty; unaccountable mood changes. IRT calibration shows a strong common dimension — all items load substantially on the first factor. When gender DIF is assessed, a systematic pattern emerges: items on the "sensitive and vulnerable" facet show negative DIF differences (women endorse more than men at the same $\theta$), while items on "nervous and tense" show positive DIF differences. These effects track the facet structure almost perfectly — they are not random item-level biases.

The interpretation is not that the items are biased in a problematic sense. Rather, the scale measures a broad stress reaction construct with genuine gender-differentiated facets. Men and women at the same level of overall stress differ in which *manifestations* of that stress they experience. Uniform DIF is the statistical fingerprint of that substantive phenomenon. Treating it as pure bias would be a mistake; understanding its source in the construct structure is the right response.

This example illustrates a general principle: **understanding DIF requires understanding construct structure**. Before concluding that DIF signals a biased item, ask whether the group difference on that item is consistent with a theoretically meaningful facet of the construct that genuinely differs between groups.

------------------------------------------------------------------------

## Discussion Questions

1.  A researcher develops a 20-item anxiety scale and reports $\alpha = .88$ in a college student sample. A clinician then uses the same scale with an inpatient psychiatric population and reports $\alpha = .91$. The researcher concludes the scale is reliable across settings. What are the two CTT-specific limitations in this reasoning, and how would IRT-based analysis address them?

2.  Two items have the same location parameter ($\beta = 0.5$) but different discrimination parameters ($\alpha = 0.8$ vs. $\alpha = 2.0$). At $\theta = 0.5$, both items have $P(\text{endorsement}) = .50$. Why would you strongly prefer the high-discrimination item for a measure designed to screen at this threshold? Quantify the difference in item information at $\theta = 0.5$.

3.  You are designing an item bank for a CAT intended to screen for clinical depression. Should you prefer items with high $\alpha$ and high $\beta$ (hard, highly discriminating items) or items with moderate $\alpha$ spread across a wide range of $\beta$ values? What does your answer depend on, and what would you need to know about the intended testing population to decide?

4.  After fitting the GRM to a polytomous anxiety measure, you find that for several items the threshold parameters $\beta_3$ and $\beta_4$ are very close together — within 0.1 $\theta$ units. What does this tell you about how respondents are using the scale, and what would you consider doing?

5.  You fit an NRM to a 10-item measure and find that for Item 6 the scoring coefficients across four categories are 0, 0.3, 0.2, and 1.9. The first three categories have nearly identical weights. What are the substantive implications of this pattern, and what would you do with this item?

6.  A measurement researcher finds large uniform DIF on several items favoring White students on a standardized achievement test. One colleague argues the items are biased and should be removed. Another argues the differences might reflect true group differences in the construct. How would you adjudicate this disagreement? What additional analyses or information would help you decide?
