---
title: "Nonlinear Regression"
author: "Michael Woller"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
standalone: true
---

```{r setup, include=FALSE}
library(ggformula)
library(DescTools)
library(dplyr)
library(MASS)

feardata <- read.csv("FearData.csv", header = TRUE)
```

## Introduction

Though it might seem like an oxymoron, linear regression can be used to model **nonlinear relationships**. The "linear" in linear regression actually refers to the fact that the model is linear in its *coefficients* (the betas), not that the relationship between the predictors and the outcome must be a straight line. This means we can include transformed predictors - such as squared terms, cubic terms, or other functions of variables - to capture curved relationships, while still estimating the model using linear regression techniques.

For example, including a term like $X^2$ allows the model to fit a curve rather than a straight line. As long as the model remains linear in the coefficients (e.g., $\beta_1 X + \beta_2 X^2$), it is still considered a linear model, even though the relationship between X and the outcome is nonlinear.

There are multiple types of nonlinear models you can run depending on what type of nonlinearity is present in the data.

### The Dataset

For this lesson we will use a **haunted house** dataset that was collected after individuals walked through a haunted house that some psychologists cooked up. Specifically, for this lesson, of interest was a "piggy" jump scare, in which someone in a pig mask jumps out with a chainsaw. The data we will look at is:

-   `FearPiggy` — how much you were afraid of the piggy jump scare (1--10 scale)
-   `EnjoyPiggy` — how much you enjoyed the piggy jump scare (1--10 scale)

We are exploring the nonlinearity of how much an individual was afraid of a "piggy jump scare" at the haunted house (`FearPiggy`) in predicting how much they enjoyed the piggy jump scare (`EnjoymentPiggy`).

It is possible that individuals only enjoyed the piggy when it was not scary or was very scary, but did not care for it when it was moderately scary. This would be a nonlinear relationship.

Note, the dataset that was given to develop the lab lesson proved to not really have any nonlinearities (spoiler), so this isn't the best example to teach with. If I ever get a hold on nonlinear data and can be bothered to come back and change some of these, I might in the future.

------------------------------------------------------------------------

## 1. Visualizing Nonlinearity

Typically it is important to visualize your data before finalizing your model, but this is especially the case when your model might be nonlinear. We need to see the type of nonlinearity present to determine what kind of nonlinear model to fit. If the relationship is a U shape, we might want a quadratic model. If it is more like a W, maybe a quartic model would be better.

We will be plotting using the `ggformula` package. Also make sure you have the `dplyr` package loaded in order to use the `%>%` piping operator later on.

```{r viz-nonlinear}
gf_point(EnjoymentPiggy ~ FearPiggy, data = feardata) +
  geom_smooth(method = "lm")
```

The scatterplot does not seem super nonlinear, but you can see how the clustering of points starts to become lower on the ends of the plot. This might be due to some trace amount of nonlinearity in a $\cap$ shape. The fitted linear line is very flat, which suggests there may not be much of a linear relationship. However, keep in mind that a flat linear line can still manifest as a curved polynomial line (think how the average of a U shape would still amount to a flat line).

To further investigate, we can plot the residuals against `FearPiggy`:

```{r residual-viz}
linear_model <- lm(EnjoymentPiggy ~ FearPiggy, data = feardata)
gf_point(resid(linear_model) ~ FearPiggy, data = feardata) +
  geom_smooth(method = "lm")
```

By looking at the left and right ends of the plot, we can see the largest deviations from the predicted line. Our straight line fits the worst at the ends, which is roughly what we took away from the previous plot. These residual plots are helpful at diagnosing at what points in the predictor range the linear line fits the worst.

*Note: the nonlinear models also do not fit very well in this dataset - it was not originally designed with a strong nonlinear relationship. The goal here is to learn the methods, not to find significant effects.*

------------------------------------------------------------------------

## 2. Polynomial Models

### What Is a Polynomial Model?

A polynomial line is a line that has one or more inflection points (a local minimum or maximum). Polynomial lines are determined by an exponential:

-   $X$ = straight line (not polynomial)
-   $X^2$ = quadratic line (one minimum or maximum, basically a U shape)
-   $X^3$ = cubic line (one minimum and one maximum)

Typically a cubic line is the highest you will go, as models with higher exponents are very rare, and you want to avoid **overfitting**. This would be building a model that fits the current dataset so closely (including random noise) that it does not generalize well to the population.

Whenever you make a polynomial line, you must include all lower-degree terms. For a quadratic model: $$Y = b_0 + b_1 X + b_2 X^2$$

For a cubic model: $$Y = b_0 + b_1 X + b_2 X^2 + b_3 X^3$$

### Fitting a Quadratic Model

You have two options. Option 1 is to create a new squared variable:

```{r quadratic-model1}
feardata$FearPiggy2 <- feardata$FearPiggy^2

polynomial_model <- lm(EnjoymentPiggy ~ FearPiggy + FearPiggy2, data = feardata)
summary(polynomial_model)
```

Option 2 is to use the `I()` function inside `lm()`, which allows you to transform a variable mid-formula:

```{r quadratic-model2}
polynomial_model2 <- lm(EnjoymentPiggy ~ FearPiggy + I(FearPiggy^2), data = feardata)
summary(polynomial_model2)
```

As we can see from these outputs, the quadratic line is not a good fit. The quadratic term is not significant ($t(99) = -1.77$, $p = .242$), so we fail to reject that there is a quadratic relationship. You can also do a $\Delta R^2$ test with and without the nonlinear term to test this more formally.

Generally, you do not interpret the polynomial term because its interpretation does not make intuitive sense. You can interpret the linear `FearPiggy` as the linear relationship (similar to what is done normally).

If you want to understand the coefficient intuitively, you can think of the $X^2$ term as an interaction, where the effect of `FearPiggy` depends on the level of `FearPiggy` itself. Knowing the polynomial coefficient is about -.05, as `FearPiggy` increases, the effect of `FearPiggy` on `EnjoymentPiggy` decreases.

### Plotting Polynomial Models

```{r plot-polynomial}
# Residuals of the polynomial model
gf_point(resid(polynomial_model) ~ FearPiggy, data = feardata)
```

```{r plot-polynomial2}
# Predicted quadratic line (don't fit a geom_smooth line here)
gf_point(predict(polynomial_model) ~ FearPiggy, data = feardata)
```

NOTE: Do not be fooled by the shape here. Note the range of the Y-axis. It is very small, meaning the quadratic line is actually very flat, which is why it is non-significant.

```{r plot-both}
# Plot predicted scores and residuals together
gf_point(resid(polynomial_model) ~ FearPiggy, 
         data = feardata, color = "red") %>%
  gf_point(predict(polynomial_model) ~ FearPiggy, 
           data = feardata, color = "blue")
```

The residuals are generally much lower than the quadratic line predicted. As I said, you can see how the quadratic line is much flatter than the unadjusted plot shows. In conclusion, a polynomial nonlinear model was a poor fit for our data.

------------------------------------------------------------------------

## 3. Piecewise Regression

### What Is Piecewise Regression?

Piecewise regression fits **separate linear relationships** across different ranges of a predictor variable by introducing one or more **breakpoints** (also called knots). Instead of assuming a single straight line describes the entire relationship, piecewise regression allows the slope to change at specified values of the predictor. This results in multiple line segments joined together at the breakpoints.

This is especially useful when the effect of a variable is not constant - for example, when a predictor has little effect up to a certain point and then becomes stronger, weaker, or changes direction. A common example is A/B clinical designs.

As a side note, piecewise regression is a type of "spline" regression. It is also related to moderation in both its interpretation and its function - the piecewise line is basically an interaction term! If you have not yet done the Moderation lesson, come back to this after.

### Segmented Piecewise Regression

A segmented piecewise model fits two disconnected line segments with different slopes and intercepts.

First, create a breakpoint indicator variable (0 for values below the breakpoint, 1 for values at or above it). We will use `FearPiggy` = 5 as the breakpoint. Always include the breakpoint in the second line segment.

```{r segmented-setup}
feardata$XJ <- (feardata$FearPiggy >= 5) # >= is greater than or equal to

# Create the interaction for the second line segment's slope
feardata$FearPiggyXJ <- (feardata$FearPiggy * feardata$XJ)
```

Then fit a model with the original predictor, the breakpoint indicator, and the interaction:

```{r segmented-model}
piecewise1 <- lm(EnjoymentPiggy ~ FearPiggy + XJ + FearPiggyXJ, data = feardata)
summary(piecewise1)
```

Looking at the output:

-   The slope of the first line segment (`FearPiggy` = .20) is not significant.
-   `FearPiggyXJ` (-.42) is the **difference** between the slope of the first and second line segments. Since this is also not significant, there is no significant difference between the two slopes.

To get the slope of the second line segment, add the two slopes together:

```{r slope-second}
-0.3829 - 0.1132
```

To get the **significance** of the second line segment slope, switch the indicator coding and rerun:

```{r segmented-model2}
feardata$XJ2          <- (feardata$FearPiggy < 5)
feardata$FearPiggyXJ2 <- (feardata$FearPiggy * feardata$XJ2)

piecewise2 <- lm(EnjoymentPiggy ~ FearPiggy + XJ2 + FearPiggyXJ2, data = feardata)
summary(piecewise2)
```

The slope of the second line segment is not significant. Note that the magnitude of the interaction is the same (.38) but the sign changes to keep the arithmetic consistent.

**Interpretation example (first segment slope):**

> For two individuals who are below a score of 5 in their fear of the piggy jump scare and differ in one unit of fear, the individual who was more afraid is expected to be about .11 point higher in their enjoyment of the piggy jump scare.

**Interpretation example (interaction):**

> The difference in slopes for individuals who have a fear piggy score of less than 5 and those who have a score of 5 or greater is about .38.

```{r segmented-plot}
gf_point(predict(piecewise1) ~ FearPiggy, data = feardata)
```

You can see the two different line segments with different projections. Don't be confused by the scale of the Y-axis. These are actually very small slopes. Adding the residuals shows the full picture:

```{r segmented-plot-full}
gf_point(resid(piecewise1) ~ FearPiggy, 
         data = feardata, color = "red") %>%
  gf_point(predict(piecewise1) ~ FearPiggy, 
           data = feardata, color = "blue")
```

### Joint Piecewise Regression

The joint piecewise regression is a more popular technique as it allows a trend to continue directly from one line to another at the knot point, like a door-hinge. Mathematically, it is two lines with different slopes but the same intercept, where one line goes to the right from the joint and one line goes left.

To make this model, create the indicator term, then **center your predictor at the joint** by subtracting the breakpoint value. This makes `FearPiggy` = 5 become 0, all values less than 5 become negative, and all values greater than 5 become positive. **DO NOT add the indicator `XJ` term** - doing so results in a segmented (disconnected) piecewise line instead.

```{r joint-setup}
feardata$XJ      <- (feardata$FearPiggy >= 5)
feardata$FearPiggyc  <- feardata$FearPiggy - 5
feardata$FearPiggycXJ <- (feardata$FearPiggyc * feardata$XJ)
```

```{r joint-model}
piecewise3 <- lm(EnjoymentPiggy ~ FearPiggyc + FearPiggycXJ, data = feardata)
summary(piecewise3)
```

-   `FearPiggyc` (.21) is the slope **before** the joint
-   `FearPiggycXJ` (-.42) is the **difference** in slopes before and after the joint

To get the slope after the joint:

```{r slope-after-joint}
-0.4191 + 0.2067
```

To get the significance of the after-joint slope, switch the indicator:

```{r joint-model2}
feardata$XJ2           <- (feardata$FearPiggy < 5)
feardata$FearPiggycXJ2 <- (feardata$FearPiggyc * feardata$XJ2)

piecewise4 <- lm(EnjoymentPiggy ~ FearPiggyc + FearPiggycXJ2, data = feardata)
summary(piecewise4)
```

The slope of the line after the joint at 5 is not significant, as we expected. The magnitude of the interaction is the same (.42) but the sign changes to keep the arithmetic consistent.

**Interpretation example (before-joint slope):**

> For two individuals who are below a score of 5 in their fear of the piggy jump scare and differ in one unit of fear, the individual who was more afraid is expected to be about .21 point higher in their enjoyment of the piggy jump scare.

**Interpretation example (interaction at joint):**

> The difference in slopes for individuals who have a fear piggy score of less than 5 and those who have a score of 5 or greater is about .41.

```{r joint-plot}
gf_point(predict(piecewise3) ~ FearPiggy, data = feardata)
```

Unlike the segmented piecewise model, this model connects when `FearPiggy` equals 5! You can imagine the Y-axis splitting at `FearPiggy` = 5, with one line going left and one going right - both meeting at the same intercept at the joint.

```{r joint-plot-full}
gf_point(resid(piecewise3) ~ FearPiggy, 
         data = feardata, color = "red") %>%
  gf_point(predict(piecewise3) ~ FearPiggy, 
           data = feardata, color = "blue")
```

The piecewise segments in blue are incredibly flat - the model is a poor fit because the slopes were not significant. If you ever wanted to do more than one joint, you can add more knots, but instead of being dummy coded like `XJ` was, you need to use sequential coding (see the Multicategorical Predictors lesson).

------------------------------------------------------------------------

## 4. Log Transformations

### What Is a Log Transformation?

Log transformations are probably the most universally applied method for dealing with nonlinearity. When you have a variable that is notably skewed, you can rescale it using logarithms, which reshapes the distribution by pulling in large values and spreading out smaller ones. This often makes relationships more linear, stabilizes variance, and improves model fit.

**NOTE:** Whenever you see "log" in statistics, it always refers to the **natural log** (otherwise denoted as "ln" in other domains). The `log()` function in R gives you the natural log by default. The `exp()` function gives you the inverse. For quick reference: `exp(log(X)) = X` and `log(exp(X)) = X`.

A caveat: the variable must be positive. If you have negative values or zeros, you need to shift your data first.

```{r log-setup}
# Shift variables up by 1 to avoid log(0) = -Infinity
feardata$FearPiggyc2    <- feardata$FearPiggy + 1 
feardata$EnjoymentPiggyc <- feardata$EnjoymentPiggy + 1 
```

### Visualizing Log Transformations

```{r log-viz}
# Non-transformed model
gf_point(EnjoymentPiggyc ~ FearPiggyc2, data = feardata) %>% gf_lm()

# Log-transformed Y
gf_point(log(EnjoymentPiggyc) ~ FearPiggyc2, data = feardata) %>% gf_lm()

# Log-transformed X
gf_point(EnjoymentPiggyc ~ log(FearPiggyc2), data = feardata) %>% gf_lm()

# Both Y and X log-transformed
gf_point(log(EnjoymentPiggyc) ~ log(FearPiggyc2), data = feardata) %>% gf_lm()
```

What you can note in the above plots is the scale of the axes. They become more condensed. Logs scale in magnitudes of distance from 0, so points further out become more compressed while points close to 0 do not get compressed as much.

```{r log-hist}
hist(feardata$EnjoymentPiggyc)
hist(log(feardata$EnjoymentPiggyc))
```

You can add slight modifications to the logarithms to adjust the severity of the transformation, but you should not really play around with this unless you know what you are doing. It is bad practice to guess numbers and transform - you **will** get critique in reviewer feedback about this.

### Log Transformation Interpretations

**Log-transformed outcome:**

When you log-transform Y, the coefficients are essentially **percentage changes** of the outcome.

```{r log-model-y}
logmodel <- lm(log(EnjoymentPiggyc) ~ FearPiggyc2, data = feardata)
summary(logmodel)
```

**Interpretation of b1:** Between two individuals who differ in FearPiggy by 1 point, the more afraid person is expected to be .01 lower in log-units of EnjoymentPiggy.

To get the exact **percent change** of the outcome: $$\%\Delta Y = (e^{b_1} - 1) \times 100$$

```{r log-percent-change}
(exp(-0.017) - 1) * 100
```

For two individuals who differ in FearPiggy by 1 point, the more afraid is expected to be 1.68% lower in EnjoymentPiggy.

**Log-transformed predictor:**

When you log-transform X, you cannot simply take the exponential of the slope to back-transform, because the slope now represents the change in Y for a one-unit change in $\log(X)$, not X itself. The effect of X depends on its current value, so there is no single back-transformed slope on the original scale.

```{r log-model-x}
logmodel2 <- lm(EnjoymentPiggyc ~ log(FearPiggyc2), data = feardata)
summary(logmodel2)
```

**Interpretation of b1:** Between two individuals who differ in $\log(\text{FearPiggy})$ by 1 unit, the individual with the higher FearPiggy is expected to be .1186 units higher in EnjoymentPiggy.

A 1-unit difference in $\log(X)$ corresponds to multiplying X by $e \approx 2.72$. To express this as a percent change:

```{r log-x-percent}
(exp(0.1186) - 1) * 100
```

For two individuals where one has about 2.72 times higher FearPiggy than the other, the person with higher FearPiggy is expected to have about 12.6% higher EnjoymentPiggy.

### Box-Cox Transformations

**Box-Cox transformations** give you a data-driven way to find the optimal transformation of the outcome to better meet regression assumptions. Instead of choosing a preset transformation (like log) by guesswork, the Box-Cox method searches across a range of possible exponential transformations and identifies the one that is most optimal for improving model fit. You can do the Box-Cox transformation using `boxcox()` from the `MASS` package.

The transformation is controlled by a parameter $\lambda$, where different values correspond to different transformations ($\lambda = 1$ is no change, $\lambda = 0$ approximates a log transformation, $\lambda = 0.5$ is a square root).

The downside is that Box-Cox transformations destroy interpretability. Once you transform your variable, the coefficients cannot be understood conceptually. For this reason, the method is not very popular since most researchers care about interpretability.

```{r boxcox}
feardata$FearPiggyc <- feardata$FearPiggy - mean(feardata$FearPiggy)
piggy_linear <- lm(EnjoymentPiggyc ~ FearPiggyc, data = feardata)

bc <- boxcox(piggy_linear, lambda = seq(-3, 3))
```

The plot shows the log-likelihoods of the model for every $\lambda$ checked. The range within the vertical dotted lines gives the best fits, with the center dotted line being the most optimal.

```{r boxcox-lambda}
best_lambda <- bc$x[which(bc$y == max(bc$y))]
best_lambda
```

```{r boxcox-model}
feardata$EnjoymentPiggyBEST <- feardata$EnjoymentPiggyc ^ best_lambda
boxcox_model <- lm(EnjoymentPiggyBEST ~ FearPiggyc, data = feardata)
summary(boxcox_model)
```

So this is the most optimal transformation, which really did not do much since the model was not nonlinear and was a poor fit in general. And now the coefficient cannot be meaningfully interpreted.

------------------------------------------------------------------------

## Well Done!

You have completed the Nonlinear Regression tutorial. Here is a summary of what was covered:

-   Visualizing nonlinearity with scatterplots and residual plots
-   Polynomial models (quadratic) using squared terms and the `I()` function
-   Piecewise regression: segmented (disconnected) and joint (connected at a knot)
-   How piecewise regression relates to interaction terms in its setup and interpretation
-   Log transformations for the outcome and predictor, including percent change interpretations
-   Box-Cox transformations as a data-driven but interpretability-destroying approach
