---
title: "Prediction"
author: "Michael Woller"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
standalone: true
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(leaps)
library(glmnet)

big5 <- read.csv("big5data.csv", header = TRUE)
```

## Introduction

Most of this course has focused on **inference**, which involves using a model to make conclusions about population-level relationships. But there is another goal in regression: **prediction**. For predictive models, we are often less concerned with how well the model fits the data we already have, and more concerned with how well it will perform on new, unseen data. A model can fit the current dataset very well simply by capturing noise (overfitting), but this does not guarantee it will generalize to future datasets.

In this lesson we will cover cross-validation methods for evaluating predictive accuracy, variable selection techniques for building optimal models, prediction intervals, and LASSO regression.

### The Dataset

For this lesson we continue with the **Big 5** personality dataset, predicting `ExistentialWellBeing` from the five personality traits:

-   `OpenMindedness`
-   `Conscientiousness`
-   `Extraversion`
-   `Agreeableness`
-   `NegativeEmotionality`

------------------------------------------------------------------------

## 1. Cross-Validation and Shrunken R

### What Is Cross-Validation?

Cross-validation is a method for evaluating how well a model is likely to perform in predicting new data. It involves splitting the data into two parts:

1.  A **training set**, used to fit the model
2.  A **testing (hold-out) set**, used to evaluate the model's predictive accuracy

The model is fit on the training data, then used to predict outcomes in the testing data. The relationship between the predicted and actual values in the testing set provides a more realistic estimate of model performance than evaluating on the same data the model was trained on.

This typically results in a smaller $R^2$ than the original model, because the model is no longer being evaluated on the same data it was trained on. This reduced value is often referred to as the **shrunken R** or $R^2$. It gives the relationship between the observed and predicted data.

### Splitting Training/Testing Data

The first thing you need to do is set your seed. This is a **very important step**. You need to be able to replicate with the exact same data splits. Your choice of seed number does not matter, but you must use one.

```{r train-test-split}
set.seed(8675309)

n           <- nrow(big5)
# Create a random index of which rows will be in the training data
train_index <- sample(1:n, size = floor(0.5 * n))

train_data <- big5[train_index, ]  # subsets big5 training data
test_data  <- big5[-train_index, ] # subsets big5 testing data
# Note: '-train_index' means all rows that DON'T include these values
```

Here we are doing a 50/50 split for demonstration purposes. In practice it is more common to use 80/20 or 90/10. Note that the results of your shrunken R will differ for simple, double, and k-fold cross-validation depending on your seed. Leave-one-out cross-validation is insensitive to the seed, as you will see below.

------------------------------------------------------------------------

### Simple Cross-Validation

This method simply trains a model with the training data, then tries to predict the testing data by funneling the testing data into the trained model. The closer the model's predictions are to the true testing data, the better the model is. We will use shrunken R to determine this.

```{r simple-cv}
# Fit the model on the training data
training_fit <- lm(ExistentialWellBeing ~ OpenMindedness + Conscientiousness 
                   + Extraversion + Agreeableness + NegativeEmotionality,
                   data = train_data)

# Predict on the testing data using "newdata ="
tested <- predict(training_fit, newdata = test_data)

# Shrunken R: correlate the testing data outcomes with the model's predictions
shrunkenR <- cor(test_data$ExistentialWellBeing, tested, use = "complete.obs")
shrunkenR
```

Our model's predictions have a moderate positive correlation (\~0.60) with actual existential well-being in new data.

```{r simple-cv-r2}
# Shrunken R^2
shrunkenR2 <- shrunkenR^2
shrunkenR2
```

About 36% of the variance in existential well-being is associated with the model's predictions in new data.

We can also compute the **out-of-sample** $R^2$, which is calculated from sums of squares rather than correlations. It asks "how much variance in $Y$ would this model explain if taken to a new sample from the same population?"

```{r oos-r2}
SSE    <- sum((test_data$ExistentialWellBeing - tested)^2)
SST    <- sum((test_data$ExistentialWellBeing - mean(test_data$ExistentialWellBeing))^2)
R2_oos <- 1 - SSE / SST
R2_oos
```

The model explains about 36% of the variance in existential well-being in the test dataset.

```{r rmse}
# RMSE: average size of prediction error (common in predictive statistics)
rmse <- sqrt(mean((test_data$ExistentialWellBeing - tested)^2))
rmse
```

On average, the model's predictions are off by about 0.59 units of existential well-being.

------------------------------------------------------------------------

### Double Cross-Validation

Double cross-validation is the same as simple cross-validation, but you train two models and test each one on the data used to train the other. This requires a 50/50 split, which is why we split the data in halves above.

```{r double-cv}
# Model 1: trained on train_data, tested on test_data (same as above)
training_fit <- lm(ExistentialWellBeing ~ OpenMindedness + Conscientiousness 
                   + Extraversion + Agreeableness + NegativeEmotionality,
                   data = train_data)

# Model 2: trained on test_data, tested on train_data (swap the halves)
second_fit <- lm(ExistentialWellBeing ~ OpenMindedness + Conscientiousness 
                 + Extraversion + Agreeableness + NegativeEmotionality,
                 data = test_data)

# Predictions
tested  <- predict(training_fit, newdata = test_data)
tested2 <- predict(second_fit,   newdata = train_data)

# Shrunken R for each model
shrunkenR  <- cor(test_data$ExistentialWellBeing,  tested,  use = "complete.obs")
shrunkenR2 <- cor(train_data$ExistentialWellBeing, tested2, use = "complete.obs")

# Final shrunken R is the average of the two
shrunkenR_dcv <- mean(c(shrunkenR, shrunkenR2))
shrunkenR_dcv

# Shrunken R^2
shrunkenR_dcv^2
```

Interpretations follow the same logic as simple cross-validation.

------------------------------------------------------------------------

### K-Fold Cross-Validation

K-fold cross-validation does the same thing as double cross-validation, but instead of doing it twice, you do it $k$ number of times. Typically $k = 10$, as there are diminishing returns when it gets much higher. This is probably the most common form of validation for regression.

```{r kfold-cv}
k  <- 5           # number of folds
N  <- nrow(big5)  # sample size
nk <- N / k       # participants per fold

# Insert fold indices into data
big5$fold <- rep(1:k, times = nk + 1)[1:N]

# Empty vector to store shrunken R per fold
shrunkenRCV_kfold <- vector(mode = "numeric", length = k)

set.seed(8675309)

# The loop does the cross-validation k times automatically
for (i in 1:k) {
  train     <- big5[big5$fold != i, ]               # training data (all folds except i)
  test      <- big5[big5$fold == i, ]               # testing data (fold i)
  modeltest <- lm(ExistentialWellBeing ~ OpenMindedness 
                  + Conscientiousness + Extraversion 
                  + Agreeableness + NegativeEmotionality,
                  data = train)
  Ypredcv   <- predict(modeltest, newdata = test)
  shrunkenRCV_kfold[i] <- cor(Ypredcv, test$ExistentialWellBeing)
}

# Average shrunken R across all k folds
shrunkenR_kfold <- mean(shrunkenRCV_kfold)
shrunkenR_kfold

# Shrunken R^2
shrunkenR_kfold^2
```

Don't worry too much about the `for` loop syntax. If you aren't familiar with creating loops in R this will be a bit too complicated for now. The key idea is that it just automates the cross-validation process $k$ times and averages the results.

------------------------------------------------------------------------

### Leave-One-Out Validation

Leave-one-out (LOO) validation evaluates a model by using all but one observation to fit the model, then testing it on the single left-out observation, repeating this for every data point in the dataset. In simple terms: "Train on almost all the data, test on one point, and repeat for every point."

Since we are testing on every single point, we don't need a seed. We start by estimating the model on all data.

```{r loo-cv}
full_model <- lm(ExistentialWellBeing ~ OpenMindedness 
                 + Conscientiousness + Extraversion  
                 + Agreeableness + NegativeEmotionality,
                 data = big5)

# Save residuals and hat values
e <- resid(full_model)
h <- hatvalues(full_model) # a measure of leverage

# Get LOO predicted value for each case
yhat_loo <- big5$ExistentialWellBeing - e / (1 - h)

# LOO Shrunken R
shrunkenR_LOO <- cor(big5$ExistentialWellBeing, yhat_loo, use = "complete.obs")
shrunkenR_LOO

# LOO Shrunken R^2
shrunkenR_LOO^2
```

------------------------------------------------------------------------

## 2. Variable Selection

Since the purpose of prediction is to have the most accurate predictions, models need to focus on **minimizing error variance** rather than getting unbiased estimates.

-   Unbiased estimates -\> Valid inference of the population
-   Minimized error variance -\> More accurate and consistent predictions of samples

Unfortunately, when doing statistics, you generally have to choose one or the other. This is the infamous **bias-variance trade-off**.

The process of minimizing error variance requires techniques called **variable selection**, which involves step-by-step algorithmic techniques to select which variables are best suited for minimizing error variance.

All an "algorithm" is is just a step-by-step guide to get a specific outcome. Machine learning methods are just automated algorithms to minimize error variance. Some are more "black box" than others, meaning the intermediate steps are unknown to us. Ridge and LASSO regression (discussed later) are examples of non-black-box automated algorithms.

The following methods are very out of date and should not really be used in practice. However, you may still encounter them in published work, so it is worth knowing how they work. In psychology, you could most likely get away with using Ridge or LASSO regression in most cases.

------------------------------------------------------------------------

### Forward Selection

Forward selection starts with no predictors and adds them one at a time. At each step, the predictor that provides the most significant improvement ($\Delta R^2$ F test) is added. The process stops when no remaining variable significantly improves the model. We use `add1()` with `test = "F"`.

Note that `~1` in `lm()` means a model with only the intercept and no predictors.

```{r forward-step1}
modelnone <- lm(ExistentialWellBeing ~ 1, data = big5)

add1(modelnone, scope = ~ OpenMindedness + Conscientiousness + Extraversion + 
       Agreeableness + NegativeEmotionality, test = "F")
```

`NegativeEmotionality` has the highest F/lowest p, so it is added first. Use `update()` with a `+` in front of the variable name to add it. Make sure every variable is in `scope =`.

```{r forward-step2}
add1(update(modelnone, ~. + NegativeEmotionality), 
     scope = ~ OpenMindedness + Conscientiousness + Extraversion
             + Agreeableness + NegativeEmotionality, test = "F")
```

`Conscientiousness` is next most significant, so it is added.

```{r forward-step3}
add1(update(modelnone, ~. + NegativeEmotionality + Conscientiousness), 
     scope = ~ OpenMindedness + Conscientiousness + Extraversion 
             + Agreeableness + NegativeEmotionality, test = "F")
```

`Agreeableness` is significant, so it is added.

```{r forward-step4}
add1(update(modelnone, ~. + NegativeEmotionality + Conscientiousness + Agreeableness), 
     scope = ~ OpenMindedness + Conscientiousness + Extraversion 
             + Agreeableness + NegativeEmotionality, test = "F")
```

No significant variables are left to add, so forward selection stops. The final model is:

`ExistentialWellBeing ~ NegativeEmotionality + Conscientiousness + Agreeableness`

------------------------------------------------------------------------

### Backward Selection

Backward selection starts with the full model and removes predictors one at a time. At each step the least useful variable (highest p-value) is dropped. The process stops when all remaining predictors are significant. We use `drop1()` with `test = "F"`. Use `update()` with a `-` in front of the variable name to remove it.

This is probably the "best" of the simple variable selection methods, though it is still not recommended to use these days.

```{r backward-step1}
full_model <- lm(ExistentialWellBeing ~ OpenMindedness + Conscientiousness 
                 + Extraversion + Agreeableness + NegativeEmotionality,
                 data = big5)

drop1(full_model, test = "F")
```

`OpenMindedness` has the highest p-value so it is dropped first.

```{r backward-step2}
drop1(update(full_model, ~ . - OpenMindedness), test = "F")
```

`Extraversion` should be dropped next.

```{r backward-step3}
drop1(update(full_model, ~ . - OpenMindedness - Extraversion), test = "F")
```

All remaining variables are significant so backward selection stops. The final model matches forward selection:

`ExistentialWellBeing ~ NegativeEmotionality + Conscientiousness + Agreeableness`

------------------------------------------------------------------------

### Stepwise Selection

Stepwise (bidirectional) selection combines forward and backward selection. Predictors are added one at a time, but after each addition the procedure also checks whether any predictors already in the model should be removed. Variables can both enter and leave the model at different steps.

```{r stepwise-step1}
modelnone <- lm(ExistentialWellBeing ~ 1, data = big5)

add1(modelnone, scope = ~ OpenMindedness + Conscientiousness + Extraversion + 
       Agreeableness + NegativeEmotionality, test = "F")
```

`NegativeEmotionality` is added first (same as forward selection).

```{r stepwise-step2}
add1(update(modelnone, ~. + NegativeEmotionality), 
     scope = ~ OpenMindedness + Conscientiousness + Extraversion 
             + Agreeableness + NegativeEmotionality, test = "F")
```

`Conscientiousness` is added. Now check whether `NegativeEmotionality` should be dropped.

```{r stepwise-drop1}
drop1(update(modelnone, ~. + NegativeEmotionality + Conscientiousness), test = "F")
```

Both are significant, so don't drop either. Continue adding.

```{r stepwise-step3}
add1(update(modelnone, ~. + NegativeEmotionality + Conscientiousness), 
     scope = ~ OpenMindedness + Conscientiousness + Extraversion 
             + Agreeableness + NegativeEmotionality, test = "F")
```

`Agreeableness` is added. Check again whether any predictor should be dropped.

```{r stepwise-drop2}
drop1(update(modelnone, ~. + NegativeEmotionality + Conscientiousness 
             + Agreeableness), test = "F")
```

All still significant. Check if any remaining variables can be added.

```{r stepwise-step4}
add1(update(modelnone, ~. + NegativeEmotionality + Conscientiousness + Agreeableness), 
     scope = ~ OpenMindedness + Conscientiousness 
             + Extraversion + Agreeableness + NegativeEmotionality, test = "F")
```

None are significant, so stepwise selection is done. The final model is:

`ExistentialWellBeing ~ NegativeEmotionality + Conscientiousness + Agreeableness`

------------------------------------------------------------------------

### All Subsets Regression

All subsets regression evaluates every possible combination of predictors and identifies the best model for each possible number of predictors, using adjusted $R^2$ as the criterion. Unlike the step-by-step methods above, this finds the globally optimal model at each size.

We use `regsubsets()` from the `leaps` package. First subset your dataset to just the variables of interest.

```{r allsubsets}
big5small <- big5 %>% select(ExistentialWellBeing, OpenMindedness, 
                              Conscientiousness, Extraversion,
                              Agreeableness, NegativeEmotionality)

subsetmodels <- regsubsets(ExistentialWellBeing ~ ., data = big5small)
summary(subsetmodels)
```

The bottom table shows the best model for each number of predictors. The number on the left is how many predictors are in the model, and `*` means that variable is included. For example, the best one-predictor model includes only `NegativeEmotionality`, and the best two-predictor model includes `NegativeEmotionality` and `Conscientiousness`.

To find the overall best model, extract the adjusted $R^2$ for each:

```{r allsubsets-adjr2}
summary(subsetmodels)$adjr2
```

The model with the highest adjusted $R^2$ wins. In this dataset the four-predictor model has the highest adjusted $R^2$, giving a final model of:

`ExistentialWellBeing ~ NegativeEmotionality + Conscientiousness + Agreeableness + Extraversion`

Note this differs slightly from forward/backward/stepwise selection, which is expected since all subsets finds the globally optimal model rather than following one sequential path.

------------------------------------------------------------------------

## 3. Prediction Intervals

Prediction intervals are used when we want to estimate not just the average outcome, but the value of a *new individual observation* given a set of predictor values. There is always uncertainty around any prediction. A **prediction interval** provides a range of values within which we expect a single future observation to fall, with a certain level of confidence (typically 95%).

It is important to distinguish prediction intervals from confidence intervals. A **confidence interval** gives a range for the *mean* outcome at a given set of predictors, whereas a **prediction interval** gives a range for an *individual* outcome. Because individual observations vary more than averages, prediction intervals are always wider than confidence intervals.

Conceptually, a prediction interval accounts for two sources of uncertainty: (1) uncertainty in estimating the regression line itself, and (2) the random error variance around that line.

If most of what we covered in this lesson is out of date, prediction intervals are the main thing you should take away. These are used everywhere and you will always need to generate some kind of interval estimate when doing anything that involves real-world prediction.

### Getting Prediction Intervals

Use `predict()` with `interval = "prediction"`. The `fit` column is the predicted value, `lwr` is the lower bound, and `upr` is the upper bound of the 95% prediction interval.

```{r pred-intervals}
full_model <- lm(ExistentialWellBeing ~ OpenMindedness + Conscientiousness 
                 + Extraversion + Agreeableness + NegativeEmotionality,
                 data = big5)

head(predict(full_model, big5, interval = "prediction")) # using head() to prevent printing hundreds of lines
```

To change the confidence level, use `level =` (default is .95):

```{r pred-interval-90}
# 90% prediction interval
head(predict(full_model, big5, interval = "prediction", level = .9))
```

For comparison, here is the 95% confidence interval per predicted score:

```{r conf-interval}
head(predict(full_model, big5, interval = "confidence"))
```

### Specific Predictions

Prediction intervals also let you estimate the uncertainty for a specific hypothetical individual. Save their predictor values in a new data frame and plug it into `predict()`.

```{r specific-pred}
newdata <- data.frame(OpenMindedness = 1, Conscientiousness = 2,
                      Extraversion = 3, Agreeableness = 0, 
                      NegativeEmotionality = 2)

predict(full_model, newdata, interval = "prediction")
```

We predict this individual has an existential well-being score of about 3, with a 95% prediction interval of \[1.86, 4.19\].

For multiple hypothetical individuals, just add more rows to the data frame:

```{r multi-pred}
newdata <- data.frame(OpenMindedness      = c(1, 0), 
                      Conscientiousness   = c(2, 1), 
                      Extraversion        = c(3, 2), 
                      Agreeableness       = c(0, 1), 
                      NegativeEmotionality = c(3, 2))

predict(full_model, newdata, interval = "prediction")
```

------------------------------------------------------------------------

## 4. LASSO Regression

### Ridge vs LASSO vs Elastic Net

Ridge and LASSO regression are more modern automated selection methods for predictive models. Instead of removing variables one at a time like forward/backward selection, these approaches keep all predictors in the model but reduce their influence based on how useful they are for prediction. This process of intentionally penalizing large coefficients is called **regularization**.

**Ridge regression** shrinks all coefficients toward zero but never fully removes any predictor from the model. It keeps everything but gives more weight to more important variables.

**LASSO regression** also shrinks coefficients, but unlike Ridge it can shrink some all the way to exactly zero, effectively removing those predictors from the model. This makes LASSO especially useful when you have many predictors or when some are not very important, because it creates a simpler, more interpretable model while reducing overfitting.

A helpful way to think about the difference: backward selection makes "yes or no" decisions about each variable, whereas Ridge and LASSO adjust how much each variable matters. LASSO can still make "drop" decisions by zeroing out coefficients, while Ridge only reduces their importance.

**Elastic Net** is a mix of LASSO and Ridge that can both shrink and remove variables. It is especially useful when predictors are highly correlated, but it is more complicated to use since you have to make additional choices about how to balance shrinkage versus removal.

In this lesson we will focus on LASSO. Note that the coefficients from LASSO should not be interpreted in the same way as standard regression coefficients. LASSO seeks to minimize prediction error, not produce unbiased estimates, so the coefficients may differ substantially from what you would get with `lm()`.

### LASSO with cv.glmnet()

To do LASSO in R, use `cv.glmnet()` from the `glmnet` package. The `cv` stands for cross-validation, which is the process for choosing the optimal $\lambda$ before LASSO is run. The predictors must be in **matrix** format.

```{r lasso}
predictors <- as.matrix(big5small %>% select(-ExistentialWellBeing))

lasso_model <- cv.glmnet(x      = predictors,
                         y      = big5$ExistentialWellBeing, 
                         family = "gaussian",  # ordinary regression
                         alpha  = 1)           # alpha = 1 specifies LASSO
# NOTE: alpha = 0 fits Ridge, alpha between 0 and 1 fits Elastic Net

lasso_model$lambda.min
```

The optimal $\lambda$ was about .02. This value does not have a direct interpretation on its own.

To visualize how prediction error changes across $\lambda$ values:

```{r lasso-plot}
plot(lasso_model)
```

The Y-axis is mean-squared prediction error (lower is better). The X-axis shows $-\log(\lambda)$, so moving right corresponds to smaller $\lambda$ (less shrinking, more complex). The numbers at the top indicate how many predictors are included at each point. The vertical dotted line marks the optimal $\lambda$.

To extract the model coefficients:

```{r lasso-coef}
# Best predictive performance
coef(lasso_model, s = "lambda.min")

# Simpler model that performs nearly as well
coef(lasso_model, s = "lambda.1se")
```

In this case they are the same, meaning allowing more complexity does not meaningfully improve prediction. LASSO kept only `NegativeEmotionality` and zeroed out everything else, which aligns with the Variable Importance lesson where `NegativeEmotionality` had by far the largest effect.

Some other useful things you can extract:

```{r lasso-extras}
lasso_model$lambda  # all lambda values tested
lasso_model$cvm     # mean cross-validation error per lambda tested
lasso_model$nzero   # number of non-zero coefficients per lambda tested
```

To get predicted scores:

```{r lasso-pred}
head(predict(lasso_model, newx = predictors, s = "lambda.min"))
```

------------------------------------------------------------------------

## Well Done!

You have completed the Prediction tutorial. Here is a summary of what was covered:

-   Cross-validation methods: simple, double, k-fold, and leave-one-out, and how to compute shrunken R and shrunken $R^2$
-   Variable selection: forward selection with `add1()`, backward selection with `drop1()`, stepwise selection, and all subsets regression with `regsubsets()`
-   Prediction intervals with `predict()` and how they differ from confidence intervals
-   LASSO regression with `cv.glmnet()` for simultaneous variable selection and coefficient shrinkage

The next lesson covers variable importance.
