---
title: "Simple Linear Regression"
author: "Michael Woller"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
standalone: true
---

```{r setup, include=FALSE}
library(ggformula)
library(dplyr)
```

## Introduction

In this lesson we will cover simple linear regression by predicting one continuous outcome from one continuous predictor. We will start with correlations, then move into fitting and interpreting regression models, extracting model information, and plotting results using both `ggformula` and base R. Note, this lesson assumes you have a basic understanding of R, though I will touch on some of R as we go. The interactable shiny app for this lesson has intro to R material in it, if you wish to review.

We will be using the built-in `faithful` dataset, which contains data on eruptions of the Old Faithful geyser in Yellowstone National Park. The dataset consists of two columns:

-   `eruptions` - the length of an Old Faithful eruption (in minutes)
-   `waiting` - the length between Old Faithful eruptions (in minutes)

------------------------------------------------------------------------

## 1. Correlations in R

### Correlations with `cor()`

Before fitting a regression model, it is often useful to look at the correlation between the predictor and the outcome. You can use the `cor()` function in R to find a correlation between two variables.

```{r correlation}
cor(faithful$eruptions, faithful$waiting)
```

This computes the Pearson correlation between `eruptions` and `waiting`. For those who are new to R, the `$` operator lets you select specific variables from a dataset.

```{r dollar-example}
faithful$eruptions
```

If your dataset contains missing values, `cor()` can fail unless you tell R how to handle them.

```{r correlation-complete}
cor(faithful$eruptions, faithful$waiting, use = "complete.obs")
```

The option `use = "complete.obs"` tells R to use only cases that have non-missing values on both variables.

You can interpret this correlation value as a standardized difference. Old Faithful eruption length are highly correlated with its waiting time between eruptions with cor = .9. In reality, the correlation just tells how many standard deviations apart are two eruption lengths.

A full and proper definition would follow as: between two eruptions of Old Faithful that differ in one standard deviation of eruption length, we expect the longer eruption length to also have a .9 standard deviation increase in wait time for the next eruption.

------------------------------------------------------------------------

## 2. Simple Linear Regression

A simple linear regression asks whether one variable (the predictor) helps us predict another variable (the outcome).

Here:

-   Outcome (Y) = `waiting`
-   Predictor (X) = `eruptions`

### Fitting Simple Linear Regression with lm()

Use `lm()` to fit a linear model ("lm" stands for "linear model"). The following code does a linear regression predicting wait time from eruptions.

```{r fit-model}
faithful_model <- lm(waiting ~ eruptions, data = faithful)
```

The `~` can be viewed as an `=` sign for the purpose of regression. The formula `waiting ~ eruptions` means: predict `waiting` from `eruptions`.

The `faithful_model` object contains much information about the regression. However, to extract the information, you need to do different things. Simply entering the model will only give you the coefficients (intercept and slope).

```{r model-print}
faithful_model
```

Typically, people use the `summary()` function to get a display of the majority of relevant model output.

```{r model-summary}
summary(faithful_model)
```

The `summary()` output gives useful model information, including:

-   Regression coefficient values (under Estimate)
-   Standard errors (under Std. Error)
-   t test statistics (under t value)
-   p values (under Pr(\>\|t\|))

For ease, significance coding (for a criteria of .05) is given in `*` asterisks. At the top of the output, you can also see the general range of residual values. At the bottom, you can see other model fit statistics:

-   Residual Standard Error (RSE) gives the typical size of prediction error
-   Multiple $R^2$ is the $R^2$ for the model
-   Adjusted $R^2$ is the adjusted-$R^2$ for the model
-   F-statistic and the values in that line is the F test for the model $R^2$

If you want confidence intervals, you need to use the `confint()` function.

```{r confint}
confint(faithful_model)
```

### Interpretations of Coefficients

Recall that interpretations need to discuss differences between two points, need to mention expected/predicted outcome values, and need to mention directionality (which is the direction of the relationship—positive or negative). Avoid within-subject causal language like "change" or "increase" because a regression model is comparing two separate data points, not changing the same point of data.

**Intercept:**

> For an Old Faithful eruption that lasted 0 minutes, the wait time is expected to be 33.47 minutes. Obviously this doesn't make sense given the context.

**Slope:**

> For two different Old Faithful eruptions that differ in their length by one minute, there is expected to be a 10.73 minute longer wait time for the longer eruption.

This is a rather finicky definition, but it is technically correct. Most people interpret regression coefficients improperly with "change" type language, which implies that we are taking an individual point and maneuvering it, but this isn't true. We are simply looking at average differences between two independent eruptions. This technicality is almost universally ignored, to the point that, practically, no one cares. But if you want to do proper psychological research you need to be aware that you are *not* making causal judgement here! That is a whole unique domain within psychological statistics.

### What Is Stored in the Model Object?

When you fit a model in R, the result is stored as an `lm` object. To check you can use `class()`, which tells you the class of any R object.

```{r class}
class(faithful_model)
```

`lm` objects contain many pieces of information about whichever regression model they represent. The `attributes()` function gives a list of everything within an object.

```{r attributes}
attributes(faithful_model)
```

Here, under `$names` (`names` being a placeholder), we can see everything that R has automatically calculated. Some especially important pieces are the following (note, I am putting each into the `head()` function to keep the output to only give me the first 10 of the outputs as opposed to hundreds of points):

```{r model-components}
faithful_model$coefficients        # regression coefficients
head(faithful_model$residuals)     # residuals
head(faithful_model$fitted.values) # predicted values
```

-   `coefficients`: the estimated intercept and slope
-   `residuals`: the difference between observed and predicted scores
-   `fitted.values`: the predicted Y values from the regression line

Recall that a predicted score is the model's estimated value of Y for each case. A residual is the difference between the observed Y value and the predicted Y value. In other words: residual = observed score - predicted score. If the residual is large, the model predicted that case poorly. If the residual is small, the model predicted that case well.

Since these two values are so universally used, there is actually shorthand for them:

```{r shorthand}
faithful_model$coef
head(faithful_model$resid)
head(faithful_model$fit)
```

### Alternative Functions for Extracting Model Information

There are also built-in functions that extract the same information.

```{r extract-functions}
coef(faithful_model)            # coefficients
head(resid(faithful_model))     # residuals
head(predict(faithful_model))   # predicted values
```

These functions are often preferred because they work across many different model objects, not just `lm()` models.

### Save Residuals and Predicted Values with transform()

One way to save new variables into the dataset is with `transform()`. This function mutates your inputted dataset with whatever new variables you specify.

```{r transform}
faithful <- transform(faithful, 
                      faithful_resid = resid(faithful_model),
                      faithful_pred  = predict(faithful_model))
```

This adds two new variables to the `faithful` dataset: `faithful_resid` = residual scores and `faithful_pred` = predicted scores. You can choose different variable names if you want.

### Save Them with the `$` Operator

Another way to save new variables is with the `$` operator. This method is very useful since it is quick. You can do this with any variable so long as its length matches the number of rows in the dataset.

```{r dollar-save}
faithful$res  <- resid(faithful_model)
faithful$pred <- predict(faithful_model)
```

This does the same thing as `transform()`, but on separate lines. The general pattern is: `dataset$new_variable <- values`.

------------------------------------------------------------------------

## 3. Plotting the Data and the Model

Graphs are extremely useful in regression because they help us see the relationship between X and Y directly. We will focus on basic scatterplots. There are many methods of plotting. We will mainly be using the `ggformula` package, but we will also show the default R function.

A scatterplot shows the relationship between two quantitative variables. Here, X-axis = `eruptions` and Y-axis = `waiting`. Each point represents one observation in the dataset.

-   If the points generally move upward from left to right, that suggests a positive relationship.
-   If they move downward from left to right, that suggests a negative relationship.

For plotting regression lines: the regression line shows the predicted value of `waiting` for each value of `eruptions`. In other words, the line is the model's best-fitting straight line through the data.

For residual plots: residuals tell us how far each observed score is from the model's predicted score. For a well-behaved linear regression model, we generally want residuals to be scattered around 0 without a strong pattern. If the residuals show a curve, funnel shape, or clustering pattern, that can suggest problems with model assumptions.

### Using `gf_point()` for Raw Data

The `gf_point()` function from the `ggformula` package gives a large variety of options for visualization. The benefit of this function is that you can conveniently specify your model within the function instead of plugging in variables manually. Though there are almost an uncountable ammount of methods you can do plotting in R.

```{r gf-scatter}
gf_point(waiting ~ eruptions, data = faithful)
```

### Using `gf_lm()` to Fit a Line

In `ggformula`, you can add the fitted regression line with `gf_lm()`. You have to first add a piping operator from the `dplyr` package at the end of the `gf_point()` function.

```{r gf-lm}
gf_point(waiting ~ eruptions, data = faithful) %>%
  gf_lm()
```

This line matches the predicted scores from the regression.

### Using `gf_point()` for Predicted Scores

Since `gf_point()` works on a selected dataset, you need your dataset to have the predicted scores saved within it.

```{r gf-pred}
gf_point(faithful_pred ~ eruptions, data = faithful)
```

### Using `gf_point()`() for Residuals

Same stipulations for the predicted scores using `gf_point()` matter here.

```{r gf-resid}
gf_point(faithful_resid ~ eruptions, data = faithful)
```

### Using `plot()` for Raw Data

Base R has the `plot()` function that can give basic scatterplots. I mainly use `plot()` for very quick visualizations to just get the idea of what data looks like. There are two methods. Manually specifying the variables using the `$` operator:

```{r base-plot}
plot(faithful$eruptions, faithful$waiting)
```

You could also do this using the `with()` function, where you basically say do `plot()` with `faithful`. Whichever method is up to you.

```{r with-plot}
with(faithful, plot(eruptions, waiting))
```

### Fitting Regression Lines on plot()

If you want to add a fitted line with `plot()`, use `abline()` with your model object in the next line. This will add the line on top of the previously generated plot.

```{r abline}
with(faithful, plot(eruptions, waiting))
abline(faithful_model)
```

### Plotting Predicted Scores and Residuals with `plot()`

There is nothing special about this. Just take wherever you saved the predicted scores and residuals and put them in `plot()`.

```{r plot-pred-resid}
plot(faithful$faithful_pred, faithful$waiting)
plot(faithful$faithful_resid, faithful$waiting)
```

------------------------------------------------------------------------

## Well Done!

You have completed the Simple Linear Regression tutorial. Here is a summary of what was covered:

-   Computing correlations with `cor()` and handling missing values with `use = "complete.obs"`
-   Fitting simple linear regression with `lm()` and reading the `summary()` output
-   Interpreting intercept and slope coefficients correctly
-   Extracting model components with `$` notation and functions like `coef()`, `resid()`, and `predict()`
-   Saving residuals and predicted values with `transform()` and the `$` operator
-   Plotting raw data, regression lines, predicted scores, and residuals with `ggformula` and base R

The next lesson covers multiple linear regression.
